Large scale prospective evaluation of co-folding across 557 Mac1-ligand complexes and three virtual screens

Version 1

Abstract

Accurate prediction of ligand-bound protein complexes and ranking them by affinity are central problems in drug discovery. While deep learning co-folding methods can help address these challenges, their evaluation has been hampered by the difficulties in assessing independence from training data and insufficiently large test sets. Here we test the ability of co-folding methods to predict the structures of 557 ligands bound to the SARS-CoV-2 NSP3 macrodomain (Mac1) that were determined after the training cut-off dates. AlphaFold3 (AF3), Boltz-2, and Chai-1 each reproduced >50% of the Mac1 ligand poses to better than 2 Å RMSD of experiment. Despite the potential for co-folding to describe protein conformational changes that stabilize ligand binding, we did not find that common conformational rearrangements, including peptide flip and a large loop opening, were recapitulated by the co-folding prediction. For AF3 and Chai-1, ligand pose prediction confidence weakly, but significantly, tracked experimental potency, while DOCK3.7 energies were only weakly correlated. Boltz-2 affinity predictions showed the strongest correlation with measured potency and, after calibration, achieved lower mean absolute error than a baseline predictor. We next assessed whether co-folding scores could rescore docking hit-lists to distinguish true ligands from non-binders among hundreds of molecules prospectively experimentally tested against AmpC β-lactamase, the dopamine D4 and the σ2 receptors. AF3 ligand pose confidence values did not separate true ligands from high-scoring false-positives as effectively as docking scores or Boltz-2 affinity predictions did. Taken together, the modest, but independent correlations of docking score and co-folding confidence or affinity suggests that integrating physics-based and deep-learning and approaches may help with hit prioritization and subsequent optimization in structure-based ligand discovery.

Introduction

Deep Learning methods such as AlphaFold2 (Jumper et al. 2021) and RosettaFold (Baek et al. 2021) revolutionized protein structure prediction. Key to their wide adoption has been their extensive testing, first in well-controlled prospective tests like CASP (Kryshtafovych et al. 2021; Ashizawa et al. 2025) and subsequently in many studies where the methods found new applications and new experimental validation (Scardino et al. 2023; Goverde et al. 2023). Notably, these methods can provide protein structural models for small molecule ligand discovery by docking/virtual screening, leveraging physics-inspired functions to score the resulting complexes (Seok and Tiwary 2025; Brocidiacono et al. 2025; Huang et al. 2006; Michino et al. 2025; Xie et al. 2025). Retrospective analyses of AlphaFold2 models suggested that the accuracy of the predicted ligand binding poses was lower than experimental structures (Karelina et al. 2023; Holcomb et al. 2023). However, prospective use of AlphaFold2 models and experimental structures (determined after the training cutoff) to identify new ligands revealed that both approaches yielded similar quality and quantity of ligands (Lyu et al. 2024; Díaz-Holguín et al. 2024). These results indicate that AF2 models may sample alternative low energy conformations that are compatible with distinct ligands from the experimentally resolved conformations, reconciling the poor retrospective recall with the prospective successes.

More recently, the advent of co-folding methods like AlphaFold3 (AF3) (Abramson et al. 2024), Chai (Boitreaud et al. 2024), and Boltz (Wohlwend et al. 2025) that leverage diffusion-based approaches for multi-component (e.g. protein and small molecule) complex predictions, have promised to extend the impact of deep learning methods from protein structure prediction to small molecule ligand-bound protein complexes (He et al. 2025). The advantages co-folding may have over traditional docking include its abilities to explore conformational variability of the protein on-the-fly and to predict ligand affinity, reportedly as well as or better than a gold-standard physics-based method, free energy perturbation (FEP) (Passaro et al. 2025). However, multiple studies have highlighted how co-folding success can track with structural similarity to training structures, consistent with substantial structural memorization (Xu et al. 2025; Škrinjar et al. 2025). An innovative approach introduced adversarial mutations to fill binding sites to show that co-folding models will still produce high confidence poses even when interactions are physically impossible (Masters et al. 2025). These results highlight pervasive memorization and hallucination in current co-folding methods and suggest that alternative computational approaches are needed to quantify the physical basis of the protein-ligand interactions.

Still, co-folding has potential for synergizing with many aspects of structure-based design and more traditional computational chemistry approaches. Two attractive applications are the prediction of ligand-bound complexes, around which structure-based ligand design is articulated (Erlanson et al. 2025); (Backus et al. 2016), and the re-scoring of large library docking hit-lists. Co-folding is orders of magnitude slower than most docking programs, which can address libraries of hundreds-of-millions to billions of compounds on CPU clusters. The method nevertheless is fast enough to re-rank virtual screening “hit-lists” and may be better at categorizing certain types of ligands (e.g. those that stabilize alternative conformations) than the simpler physics-based methods. Moreover, the errors and biases from co-folding affinity estimation methods (Passaro et al. 2025) may be independent from those in docking, suggesting that each method may have different strengths in distinguishing likely from unlikely ligands. However, partly because of the lack of widely adopted community prospective tests, akin to CASP (Gathiaka et al. 2016) and partly because of the challenges of developing, testing, and potentially optimizing new chemical matter for ligand activity studies (Ackloo et al. 2022), co-folding methods have not yet been subject to the rigorous, large-scale testing that helped establish the utility, domain of applicability, and ultimately optimization of the early protein structure-prediction methods.

It therefore seemed interesting to test co-folding predictions of the co-complexed structures and affinities of ligands in a drug-discovery series from outside of their training sets, and to test their ability to distinguish likely from unlikely ligands from large library virtual screens (Cabeza de Vaca et al. 2025). Here, we evaluated the performance of three widely-used co-folding methods, AF3, Chai-1, and Boltz-2 against two benchmarks that speak to domain utility and are largely outside the benchmarks against which they trained. First, we tested the ability of the co-folding methods to recapitulate the crystallographic structures of 557 different ligands bound to the SARS-CoV-2 Macrodomain (Mac1), an emerging antiviral target (Suryawanshi et al. 2025). This dataset represents the scale and depth typical of the early phase of industrial structure-based drug design campaigns. Second, we investigated the ability of the co-folding methods to distinguish true- from false-positives in hit lists from recent large library, prospective docking against three well-behaved targets: AmpC β-lactamase, the σ2 receptor, and the dopamine D4 receptor (Lyu et al. 2019; Alon et al. 2021; Liu et al. 2025). In the campaigns against these targets, between 500 and 1500 molecules were prospectively tested across a range of docking ranks identifying ligands with previously unexplored chemotypes as well as finding many docking false positives. Contemporaneous work has highlighted the importance of using AF3 in the same experimental datasets from ultra-large library docking campaigns (Menon et al. 2025). Our analyses reveal the strengths and limitations of different co-folding methods versus traditional docking and point to potential synergies that can improve probe and lead discovery pipelines.

Results

A set of 557 previously unreported Mac1 ligands complexes

As part of a drug discovery effort (Suryawanshi et al. 2025; Jaishankar et al. 2025; Correy et al. 2025; Flowers et al. 2025; Gahbauer et al. 2023; Schuller et al. 2021), we tested thousands of novel small molecules as inhibitors of Mac1, a new target for SARS-CoV2 and other viruses. This led to potent inhibitors that are active in vivo as anti-virals (Suryawanshi et al. 2025; Jaishankar et al. 2025). As part of this work, we determined x-ray crystal structures for ~1000 small molecules bound to Mac1, typically at close to 1 Å resolution, but only a limited number of these were disclosed in the papers describing the more advanced candidates. Moreover, only up to 326 of our molecules, largely consisting of initial fragments and early hit-to-lead material, were available in public databases prior to the training cutoffs for the co-folding methods (e.g. 2021-09-30 for AF3, 2023-06-30 for Boltz-2). Apart from their role in driving inhibitor optimization for Mac1, we thought that these structures might be a resource for evaluating computational approaches against a set that is representative of the scale and depth often seen in industrial structure-based drug design campaigns.

From our internal dataset, we identified 557 ligands where we determined X-ray structures that have not been disclosed previously (Fig. 1a, STable 1). These molecules all bind in the active site where they compete with the substrate, ADP-ribosylated (ADP-r) peptides. There are several stereotyped interactions across this set, which have been described in previous fragment and lead optimization efforts (Gahbauer et al. 2023; Schuller et al. 2021; Correy et al. 2025), including: interactions with backbone NH groups of Phe156 and Asp157 (the oxyanion site), aromatic packing against the side chain of Phe156 (mimicking the adenosine ring of ADP-r), and hydrogen bonds to Asp22 and Ile23 (the upper active site) (Fig. 1a). Because these ligands were synthesized during an active campaign and many did not meet evolving thresholds for single point % inhibition, only 202 have fully measured dose response IC50 estimations by a HTRF-based peptide displacement assay. For these compounds, the measured potencies range from sub-micromolar to >500 uM (Fig. 1b). This distribution is skewed towards weaker affinities as more potent molecules were progressed and disclosed in papers describing in vivo-active lead molecules (Suryawanshi et al. 2025; Jaishankar et al. 2025).

Structure diversity of 557 previously unreported ligands co-folded with the SARS-CoV-2 Mac1 (NSP3 macrodomain 1).

a) Global superposition of all 557 ligands within the Mac1 binding pocket highlighting residues involved in key interactions. b) The affinity distribution of 202 molecules determined by HTRF (drug-response curves at 8 concentration points). c) Principal Component Analysis on pairwise MCS% and Tc to illustrate the chemical diversity of new molecules in the space. d) The highest MCS% vs. ECFP-4 Tc values in the training set for each Mac1 compound. At right are examples of the low similarity, scaffold hop, and high similarity distributions, respectively.

The diversity of the 557 ligands is important for understanding the performance of the co-folding methods. By topological similarity, using ECFP4-based pairwise Tanimoto coefficients (Tc), the ligands were diverse, with pairwise Tc clustering between 0.2 to 0.4 (SFig.1a, STable 2). For context, a scaffold-hop for the ECFP4-based fingerprint is thought to occur at Tc values below 0.35 (Cramer et al. 2004; Muchmore et al. 2008), while random similarity occurs around Tc values of 0.25 (Hert et al. 2008). As the Tc metric has well-known liabilities that can mask what a chemist would see as similarity, we also calculated pairwise Maximum Common Substructures (MCS) and found a distribution that centers around 44.8% (SFig.1b, STable 2). Clustering based on pairwise MCS > 35% revealed only four distinct cluster heads and a limited number of outliers (Fig. 1c, STable 2, SFig.2). To quantify the potential for model performance to be driven by memorization, for each ligand in our set, we retrieved the most similar ligand that was deposited prior to the training cutoff date by both MCS and Tc (Fig. 1d). Here, the Tcs have a trimodal distribution with a small number having Tc > 0.4 (indicating relatively high similarity for this fingerprint), and two larger distributions centered around 0.35 (indicating scaffold hops) and 0.2 (indicating low similarity to any ligand). The MCS similarities spanned a broad range centered around ~60%, consistent with the majority of the dataset representing a small number of conserved scaffolds bearing diverse peripheral substitutions.

Co-folding can accurately reproduce poses of ligands dissimilar to those trained

Next we assessed how well co-folding methods could predict the structures of the 557 complexes. Three co-folding methods (AlphaFold3, Chai-1, and Boltz-2) were given the SMILES strings of each ligand and the sequence of the enzyme as inputs. In parallel, we performed classic docking with DOCK3.7 using the SMILES of the ligand and the three dimensional structure of the enzyme bound to an early inhibitor (PDB: 5SQW, (Gahbauer et al. 2023) that is roughly contemporaneous with the ligands in this datasets. To assess whether ligands were correctly predicted in the right area of the active site (Nittinger et al. 2025), we used a threshold of ligand center of mass < 2.5 Å after alignment of protein structures: 519 out of 550 AlphaFold3 poses (94%), 509 out of 551 Chai-1 poses (92%) and 511 out of 548 Boltz-2 poses (93%). With DOCK3.7, only 83% (450 out of 544 poses) passed this cutoff. In addition, roughly 1.5% of 557 co-folded or docked molecules did not pass the RDkit-based sanitization pipeline necessary for RMSD comparisons. Therefore, the total number of poses compared for prospective prediction differed slightly among all co-folding and docking methods (STable 3).

Across the co-folding methods, AF3 and Chai-1 performed qualitatively similar with a roughly trimodal distribution centered around ~1 Å (the mode), 2 Å and 4 Å RMSD values (Fig. 2a). Using the field standard, 2 Å cutoff for correct predictions, AF3 and Chai-1 predicted ~70% of the ligands correctly. Boltz-2 performed less well, with more structures shifting to the 2 Å distribution, and an overall predictive accuracy of 52%. This result was surprising because the later training cut-off date for Boltz-2 meant that it could potentially benefit from additional Mac1 ligand-bound structures. The distribution for docking was more unimodal centered around an RMSD of 2Å, yielding a predictive accuracy of 41%. At least for this set, the co-folding models outperformed DOCK3.7 in accurately predicting the poses of bound ligands.

Co-folding outperforms DOCK3.7 in predicting the ligand pose for > 500 ligands against Mac1 binding site.

a) Histogram distributions of ligand heavy-atom RMSD (L-RMSD) between predicted and crystallographic poses obtained with AF3, Chai-1, Boltz-2 and DOCK3.7, and the success rate defined as RMSD < 2 Å, b) Demonstration of alternate conformations of Mac1 binding pocket (twisted backbone and open state) and ability of AF3 to capture these conformational changes (residues in green are from PDB ID: 5SQW, and residues in pink have altered conformations), c) AF3 L-RMSD pose recovery is compared with TC ECFP4 and MCS% and accurate poses for test molecules with poor similarity to the closest molecule in the training set.

Next we investigated the potential for co-folding methods to identify alternative conformations of the protein that are stabilized by ligand binding. This application is a potentially large differentiator relative to virtual screening, which generally uses a fixed structure. Although protein alignment gave generally low global C-alpha RMSDs (range 0.1 Å – 0.4 Å; STable 3), rare local and high magnitude conformational changes observed experimentally were not captured by co-folding methods. Previously published structures indicated two characteristic states with significant conformational changes: a twisted state, characterized by a peptide backbone flip around Phe156, and an open state, characterized by a ~8 Å loop rearrangement including Gly130 and Ala129 (Fig. 2b, SFig.3, STable 4). The open state has also been observed in the human MacroD1 enzyme (PDB 2X47, (Chen et al. 2011) and PDB 6LH4, (Yang et al. 2020)), suggesting it might be a common feature across macrodomains. We observed that only 4/19 twisted states and 0/20 open states were correctly predicted by AF3. Interestingly, AF3 predicted 8 false positive twisted states, where the ligand was experimentally observed in the normal ground state. This result suggests that there is only a limited ability to predict conformational adaptations, which admittedly may be, at least in part, stabilized by crystal contacts. We next asked how often the predictions recapitulated hydrogen bonding interactions in the Mac1 active site (SFig.3a). Among the dominant interactions, we found a low false positive rate for hydrogen bonds with D22 and I23 at the top of the active site, and mispredictions in both directions for hydrogen bonds with F156 and D157 at the oxyanion site (SFig.3b). Thus, despite failing to predict large conformational changes and aspects of the detailed hydrogen bonding interactions, AF3 still correctly placed these ligands with low RMSDs relative to the crystal structure pose (Errington et al. 2025).

Finally, we asked how much of co-folding success might be explained by memorization of Mac1 ligands deposited prior to the training date (Fig. 1d). We measured the correlation between AF3 ligand RMSD, and either the Tc (ECFP4) or %MCS of the most similar ligand within the training set, finding no meaningful signal (Fig. 2c). Examples illustrated have low similarity in both metrics, and mac-x5015, mac-x3538 are not from the commonly shared purine scaffold. Nonetheless, even for these cases, AF3 co-folded poses are recovered to within 1 Å accuracy. The weak correlations between RMSD and both similarity metrics illustrate the ability to predict binding modes for ligands dissimilar from the training set.

Model confidence and energy scores reflect pose reproduction and affinities

A central question for any co-folding or docking method is whether its internal confidence or energy scores track with experimental pose accuracy and affinity. We used the structure confidence scores reported by AF3, the per atom Predicted Local Distance Difference Test (pLDDT) scores for ligand atoms, and for Chai-1 the interface predictive template modeling (ipTM) score. While these metrics are intended to estimate the model confidence in accuracy, they are not explicitly designed to predict affinity. In contrast, the Boltz-2 affinity module predicts IC50 (pIC50) which should correlate with affinity, whilst the DOCK3.7 score, though in kcal/mol, is not expected to correlate strongly with experimental affinity (Shoichet 2004; Irwin and Shoichet 2016; Tirado-Rives and Jorgensen 2006). Boltz-2 affinity score and DOCK energies do not explicitly estimate the accuracy (RMSD) of the prediction.

We first generated Receiver Operating Characteristic (ROC) curves to measure the retrieval of True Positives (good score, RMSD < 2Å) versus False Positives (good score, RMSD > 2Å) (Fig. 3a). We observed the best performance for AF3 L-pLDDT (AUC=0.76) and Chai-1 ipTM (AUC=0.73). This indicates that the internal model confidence has predictive power and, indeed, we observed that these scores are correlated with RMSD (SFig.4). Perhaps not surprisingly, the performance of the DOCK3.7 energy (AUC=0.67) and Boltz-2 pIC50 (AUC=0.55) scores, which are designed to predict affinity, was poorer than those designed to predict accuracy. The DOCK3.7 scores correlated with Mac1 ligand RMSD with a Pearson R of 0.35 (p < 0.001), which is in line with expectations from historical analyses of docking studies (Wierbowski et al. 2020). However, Boltz-2 affinity scores did not correlate with the predicted pose (Pearson r = −0.03, p = 0.49), which likely results from the fact that its internal structure prediction confidence module is separated from the affinity module (Passaro et al. 2025). Indeed, using the Boltz-2 structure prediction confidence metric (iPTM), yields an AUC of 0.73, comparable to AF3 and Chai-1. Since confidence scores correlated with RMSD, we also investigated how each method was able to predict poses and scores depending on similarity of molecules to the trained set (SFig.5). As before (Fig. 2c), the performance of co-folding and docking scores did not show a strong dependence on novelty relative to the training set. This result suggests that co-folding can accurately reproduce poses of ligands dissimilar to those on which it trained, and that the model internal confidence scores reflect how well poses are reproduced.

Complementarity between co-folding confidence metrics and ligand-pose accuracy.

a) Receiver Operating Characteristic (ROC) curves for affinity/energy and confidence scores in AlphaFold3, Chai-1 and Boltz-2, with the success rate being defined as L-RMSD < 2 Å. b) Chemical structure of a compound that was challenging to model with all docking, co-folding methods (mac-x4091), and one that was well-modelled with all methods (mac-x3927), c) Overlay of co-folded (AlphaFold3, Chai-1, Boltz-2) and docked ligand poses on ground-truth, for hits with similar IC50, but different structures. Hydrogen bonds with co-folded, docked ligands are shown in red, and those with crystal ligands are shown in yellow.

Representative examples of correct and incorrect pose predictions across co-folding and docking methods are illustrated with mac-x3927 and mac-x4091, respectively (Fig. 3b,c). Both compounds exhibited comparable experimental IC₅₀ values (20 µM for mac-x4091 and 42 µM for mac-x3927) and similarity to molecules in the training set (Fig. 3b). The molecule consistently predicted correctly (mac-x3927) was a small, rigid fragment comprising a purine scaffold rigidly linked to a fluorobenzene moiety, favoring a well-defined binding mode. In contrast, the compound mispredicted by all methods (mac-x4091) was larger and more conformationally flexible, with an alkyne linker prone to flipping, potentially altering the orientation of the fluorobenzyl ether substituent. Although this group packs against a crystal contact, the high predictive confidence/affinity scores across co-folding methods (AF3 L-pLDDT = 86.4, Chai-1 ipTM = 0.91) are likely explained by the adenosine mimic purine group where predictions recapitulate key interactions observed in the experimental structure.

To understand if prediction errors between methods are correlated, we directly compared the agreement between AF3 to Chai-1, Boltz-2 or DOCK3.7 predictions with the agreement between AF3 and experimental ground truth (SFig.6). We observed that while there are a small number of structures where AF3 predicts much better than other methods (high AF3-other prediction RMSD, low AF3-experiment RMSD), most large errors are either missed in a similar way (low AF3-other prediction RMSD, but high AF3-experiment RMSD) or are both extreme mispredictions (high AF3-other prediction RMSD and AF3-experiment RMSD). Consistent with this result, in comparing RMSD relative to experimental structures for all methods (SFig.7), we observed that Chai-1 and AF3 exhibited the strongest agreement (r = 0.72), whereas poses from DOCK and Boltz-2 diverged more (r = 0.45 for AF3 vs. DOCK, and r = 0.52 for AF3 vs. Boltz-2). Collectively, these results suggest that the co-folding methods (AF3, Chai-1) may be complementary to more physics-based approaches like DOCK3.7 as the errors are less correlated than observed for the different cofolding methods.

Second, we asked how well these scores correlate with experimentally measured potency (Fig. 4a). The correlation was best for the Boltz-2 pIC50 (r = 0.6), significantly outperforming DOCK when tested with Friedman and Conover tests with Bonferroni corrections (p = 0.0057). Prior to any linear correction, the mean absolute error (MAE) in affinity prediction was ~1.6 kcals (1.19 pIC units), with most affinities predicted as too potent. Empirically correcting with a linear fit reduced this error to ~0.7 kcals (0.54 pIC units). To put these results in context, simply predicting the mean pIC50 from the dataset for every compound would yield a baseline MAE of 0.93 kcal because of the relatively limited dynamic range of the dataset. The DOCK3.7 energy correlated more weakly with potency (r = −0.225), consistent with previous studies that have observed only weak correlations with affinity from virtual screening energies (Kinnings et al. 2011; Jones et al. 2021). The AF3 co-folding confidence metrics correlated slightly better than docking with potency (r = 0.314, Fig. 4a), but the post-hoc Conover tests with Bonferroni corrections showed the difference was not statistically meaningful. Despite this, a correlation between pose accuracy and potency was more relevant with AF3 (r = −0.297, Fig. 4b), as more accurate poses tend to be more confidently predicted (SFig.4). In contrast, there was no correlation between docking accuracy and potency (Fig. 4b). Boltz pIC50 had the highest correlation between pose accuracy and affinity (r = −0.387). This result is interesting because in Boltz-2 affinity prediction and confidence are decoupled architecturally and have little correlation with each other across the Mac1 dataset (SFig.4).

Co-folding scores and pose accuracy correlate with Mac1 binding affinity for 202 newly synthesized Mac1 ligands.

a) AF3 ligand pLDDT (L-pLDDT) and Boltz-2 predicted pIC50 correlate with experimental pIC50 more strongly than DOCK energy scores, b) Lower co-folded ligand RMSDs associate with higher affinity, while docked pose RMSDs do not. Pearson’s Correlation Coefficients and Mean Absolute Errors (MAE) are shown. The baseline MAE was 0.68 pIC (or 0.93 kcals), measured against a fixed average pIC50 (pIC50 = 4.95). Friedman tests with Conover post-hoc comparisons and Bonferroni corrections are used to compare MAE between methods.

Collectively, these results suggest that the categorization abilities of the co-folding confidence metrics, and their correlation with affinity, could mark an important advance. Still, the Mac1 set, as large as it is, represents only a single system with ligands that share a limited set of pharmacophores. We wanted to broaden testing to more targets and more diverse ligands, such as those encountered in virtual screening hit lists with their focus on new chemotype discovery.

Co-folding struggles to reliably re-rank Docking Hit-lists

Encouraged by the ability of co-folding to reproduce ligand structures and affinities, we were interested to see if the methods could also improve hit identification from large library docking screens. In these screens, hundreds of millions to billions of molecules are docked against a target and scored for complementarity to a binding site. High ranking molecules are then synthesized and tested for binding. While large library docking has successfully discovered potent novel chemotypes in recent studies (Lyu et al. 2019; Gorgulla et al. 2020; Stein et al. 2020; Sadybekov et al. 2022; Fink et al. 2022; Singh et al. 2023; Liu et al. 2025), the method is plagued by false positives (molecules that rank well but are inactive). We reasoned that if co-folding can predict the structures and activities of true ligands confidently, perhaps it could distinguish docking true binders from false positives.

To test this idea, we analyzed experimentally tested compounds from three large and heavily tested docking campaigns: 506 experimentally-tested compounds from a 490-million molecule screen against the σ2 receptor (Alon et al. 2021), 1,293 tested compounds from a 1.7-billion molecule screen against AmpC β-lactamase (Liu et al. 2025), and 541 tested compounds from a 138-million molecule screen against dopamine D4 receptors (Lyu et al. 2019) with AF3 (the best performing model from the Mac1 set), DOCK scoring function, and the Boltz-2 pIC50 score. These test sets differ from the Mac1 benchmark in three ways. First, they are a mix of molecules that would turn out to be true binders (247 in AmpC, 201 in σ2 and 205 in Dopamine D4) and high-ranking docking false positives as judged by measured potencies (worse than 400 μM in AmpC, worse than 2.0 μM against the σ2 receptor, and worse than 7.8 μM against the dopamine D4 receptor). For these molecules, the challenge is not only predicting the bound structures of true ligands, but distinguishing true ligands from decoy molecules that by many biophysical criteria fit the binding site well. Second, the docking molecules are more diverse than the Mac1 set (SFig.2), which largely falls into four substructure clusters representing conserved pharmacophores. The docking sets were selected, in the original studies, for their topological diversity, and their dissimilarity to previously known ligands. For each target, the ligands group into more than 10-times the number of clusters than the Mac1 dataset, consistent with a greater diversity. Indeed, the average pairwise ECFP4 Tc values within the three sets vary from 0.12 to 0.15, all well-below scaffold-hop cut-offs for this fingerprint (SFig.2, STable 2). Finally, the sets were all selected for novelty and there was little similarity between the ligands in the hit-lists and known ligands, which ranged between 0.11 and 0.19 by Tc, and between 33 to 42 by MCS%. These ranges were lower than our prior Mac1 dataset (where Tc is distributed around 0.36, and MCS% is distributed around 60.1%), and the lack of similarity with the training set for target systems with hit-lists could be explained by having relatively fewer ligand structures to compare to in the PDB (SFig.8, STable6). These conditions, while representing new and difficult tests for co-folding, are the common case for structure-based ligand discovery campaigns. For validation, there are more measured IC50 values, but they are admittedly not as deeply supported by structural information as the Mac1 set.

We observed the strongest discrimination between active and inactive molecules in the σ2 dataset, which contains 201 confirmed actives (apparent Ki values between 2.5 nM and 1.5 µM) and 305 inactive false positives (Fig. 5a). DOCK3.7 scores well-separated binders from non-binders (median –55.0 vs –42.5 kcal mol−1; AUC = 78.8), though the correlation between docking score and pKi for true ligands was relatively weak (r = –0.211). AF3 L-pLDDT showed a poorer ability to discriminate between true ligands and false positives (median 75.9 vs 74.1, AUC = 56.1). Intriguingly, despite the worse discriminatory ability, the AF3 correlation with binders pKi was not much worse than docking’s (r = 0.183). Against the σ2 set, Boltz-2 pIC50 performed better still, with high discriminatory power (median 6.70 vs 5.38, AUC = 83.8) and the strongest correlation of affinity predictions against true positives (r = 0.486). Among active molecules, Boltz-2 pIC50 had a MAE of only 0.7 kcal and a linear calibration reduced that error to 0.54 kcal (SFig.9). This absolute estimation of IC50 was superior to its performance by ~0.5 kcal on the Mac1 dataset; however, it is restricted only to true positives and the performance could be much worse if the false positives were accounted for in the comparison. Since inactive predictions were in the same pIC50 range as active molecules, as an optimistic limit of the performance that includes inactive molecules, we assigned an IC50 of 4 µM, which is 2X our threshold for declaring a non–binder, to all false positives and calculated the MAE. With this assumption, the MAE is 0.92 kcal (0.56 kcal after linear calibration). A less generous assumption would be to randomly assign an IC50 between the pKi threshold (2 µM) and pKi = 1 (100 mM). Under this assumption, the MAE is 1.66 kcal (1.55 kcal after linear calibration) (SFig.10), which is in line with previously published estimates of the accuracy of this method (Passaro et al. 2025) and only a slight improvement over the MAE (1.84 kcal) produced by predicting the dataset average affinity for each datapoint.

Classification of known active hits from docked false positives from an initially-screened docked molecules.

The leftmost panel shows affinity scores from Boltz-2 (Boltz-2 pIC50), the middle panel for AF3 L-pLDDT and the right panel for absolute value of DOCK energies (kcal/mol) from DOCK3.7. Docked hit lists from a) σ2, b) Dopamine D4, c) AmpC are grouped into known hits and non-hits, and score distributions are shown in violin plots. Komolgorov-Smirnov test and its KS-statistic (KS stat) is used to indicate meaningful separation of two distribution curves. Asterisks indicate the p-value of KS statistic < 0.001. ROC curves show an ability to distinguish known actives from the docked list with different scoring methods.

For AmpC (Fig. 5b), a total of 1,293 high-ranking molecules were experimentally tested for inhibition (Liu et al. 2025). Among these were 247 molecules that turned out to be true inhibitors. We observed that docking (AUC = 76.5, median −85.6 vs −82.8 kcal/mol) was much stronger than Boltz-2 (AUC = 68.1, median 5.16 vs 4.87) or AF3 (AUC = 60.5, median 76.6 vs 72.0) in discriminating actives from inactives. The correlation of metrics to pKi among actives was relatively weak for DOCK3.7, AF3 and Boltz-2 in this case. The Boltz-2 pIC50 values for AmpC were particularly high in absolute error 1.81 kcal; however, linear correction reduced this error to only 0.42 kcal. Including inactive molecules degraded performance to 2.33 kcal (0.32 kcal with linear correction) under the optimistic and 3.29 kcal (1.04 kcal with linear correction) under the random IC50 assumptions (SFig.10). Similarly, this represents a small reduction relative to the MAE (1.07 kcal) produced by predicting the dataset average affinity for each datapoint

The dopamine receptor also proved to be more challenging than σ2, particularly for AF3. With D4 with 205 confirmed actives (apparent Ki values between 3.1 nM and 7.8 µM) and 336 inactive false positives with, we observed that actives were distinguished from docked false positives by docking (−63.9 vs −52.0 kcal/mol) and Boltz-2 (6.15 vs 5.72) scores, both with AUC ~ 71. In contrast, AF3 L-pLDDT values showed little separation (65.4 vs 66.2) and an AUC = 46.4, which is slightly worse than random (Fig. 5c). Among active molecules, the docking correlation was slightly stronger for D4 than it was for σ2 (r = −0.247), but performance degraded for AF3 (r = −0.109, notably a positive correlation is expected) and Boltz-2 pIC50 (r = 0.284). Despite this reduced linear correlation for the Boltz-2 pIC50, the MAE was 0.89 kcal (0.55 kcal after correction). The performance degraded to 1.15 kcal (0.57 kcal after linear correction) under the optimistic assumption of 2X IC50 for all false positives and to 1.97 kcal (1.4 kcal after linear correction) under the randomly assigned assumption (SFig.10). This result is also a small improvement over the MAE (1.46 kcal) produced by predicting the dataset average affinity for each datapoint

An alternative way of looking at how well the scores can identify active ligands is to perform a hit enrichment analysis using normalized ranking scores (pProp, the negative logarithm of the percentage rank (Liu et al. 2025): a molecule that ranks among the top 0.1% of the docked library would have a pProp of 3, one that ranked among the top 0.01% of the docked library would have a pProp of 4, and so forth). pProp assigned here is based on the ranking within the docked list, for direct comparison between scoring methods and largely confirmed the trends above (SFig.11). Both DOCK3.7 and Boltz-2 clearly enriched actives across most targets, with hit rates plateauing at 40 - 50% for high-ranking compounds. AF3 L-pLDDT showed meaningful and comparable enrichment only for AmpC, while providing no enrichment signal for σ2 and D4. Overall, the success of the co-folding methods at predicting structures and correlating with affinities among true Mac1 inhibitors was far less apparent in their ability to distinguish true- from false-positives among docking hit-lists, and to correlate with affinities among these highly diverse molecules. Indeed, the methods showed little advantage over docking, a method that is about four orders of magnitude faster. An important caveat is that the hit-lists were composed of molecules prioritized by docking in the first place, giving it an advantage on these particular sets.

Discussion

A striking observation from this study was the ability of the co-folding methods to predict ligand structures across 557 previously unpublished Mac1 complexes, and for the co-folding confidence metrics to correlate with ligand affinities. By both criteria all co-folding methods were meaningfully better than docking, at least as represented by DOCK3.7. Affinity prediction by Boltz-2 also performed well with mean absolute error that was ~0.2kcal better than a baseline expectation of simply using the mean IC50 value. This small improvement was a trend that also held up in the large docking sets as well. Conversely, when confronted with high-ranking molecules from large-library docking, a diverse mix of true- and false-positives with thousands of novel molecules prospectively tested against three unrelated targets, co-folding was not meaningfully better than docking at ligand categorization or at affinity prediction, indeed often worse. This despite the four log-orders greater speed of the docking and its well-known approximations and errors, long a topic of regret in the community (Jorgensen and Tirado-Rives 2005; Irwin and Shoichet 2016). These apparently contradictory results may be reconciled by the different tasks of predicting geometries for molecules that were all true ligands versus that of categorizing binders and non-binders within a highly diverse docking hit-list intentionally dissimilar to known ligands.

It may be that success and relative failure in these different tasks points also to a complementarity between physics-based docking and co-folding in the related, but different tasks of ligand discovery and optimization. The weak correlation between the methods (SFig.12) further suggests complementarity between deep learning and physics based approaches. Docking and related methods are designed for categorization among diverse, novel sets of candidate molecules, drawn from the vast new make-on-demand space (Gentile et al. 2020; Radaeva et al. 2023; Sadybekov et al. 2022; Sadybekov and Katritch 2023; Gorgulla et al. 2021; Lyu et al. 2019, 2023; Liu et al. 2024, 2025), whereas the co-folding methods, relying on learned patterns, may be relatively less effective. Where they have complementary strengths is in predicting the accurate geometries of ligands, once known, and their ranking within a series, both of which are crucial for affinity maturation once a series is discovered and with both of which docking has historically struggled. It is in these two linked but different aspects of ligand discovery and ligand optimization where the true complementarity between the methods may lie, at least for now.

Several caveats merit airing. The Mac1 benchmark, while diverse by ECPP4 Tc, is nevertheless dominated by four groups when clustered by MCS and benefits from a large training set of fragment molecules bound to the protein in the training set (Schuller et al. 2021). This may inflate apparent pose recovery rates, although Boltz-2 predicted poses were poorly recapitulated even with its much later training cutoff dates. Unexpectedly, even when the co-folding methods got important conformations of the enzyme binding site wrong, the ligands were still often posed correctly, speaking to a non-physical aspect of the ML methods that others have also described (Masters et al. 2025; Škrinjar et al. 2025). Even though the prospective docking hit lists included thousands of diverse molecules, they still represent only three targets, and so offer a limited view of molecular recognition. Finally, comparing co-folding to docking based on hit-lists themselves selected by docking is arguably unfair to co-folding. Counter-balancing this is the inclusion, in each of the three hit lists, of molecules that had mediocre and poor docking scores intentionally selected to test the correlation between docking score and hit-rate. Here too, the correlation between co-folding score and likelihood to bind, what we sometimes call a “dock-response-curve” was no better than docking’s, often worse (SFig.11).

These caveats should not obscure the central lessons from this study. Against a benchmark of 557 topologically diverse Mac1 ligand-bound structures, determined to true atomic resolution and largely dissimilar to structures already in the PDB, each of the co-folding methods predicted ligand poses and interactions with high-fidelity to the experimental structure. For Boltz-2, there was a strong and highly significant correlation between predicted affinity and experimental IC50 (compared to a modest correlation for AF3 L-pLDDT and experimental IC50). In both cases, co-folding performance was much better than that of DOCK3.7, although the correlation was only significantly different for Boltz-2 pIC50 versus experimental pIC50. Conversely, when the more homogeneous benchmark of true ligands was replaced by a diverse set of docked molecules, composed of what would turn out to be hundreds of true ligands and thousands of false positives, the co-folding methods did no better than docking and often did worse. These observations are supported by a fascinating study on some of the same ligand sets as investigated here, using AlphaFold3, reaching similar conclusions (Menon et al. 2025). The relatively strong performance by docking is surprising given its well-known weaknesses and its much reduced computational cost (and therefore greater speed). What emerges are approaches that may be complementary in their roles in ligand discovery (docking) and structure-based ligand optimization (co-folding), areas that will merit future research and application.

Methods

Mac1 X-ray crystallography

Ligands were soaked into SARS-CoV-2 Mac1 crystals (P43 crystal form (Schuller et al. 2021)), and data collected and reduced, as described previously (Gahbauer et al. 2023; Correy et al. 2025). We modeled ligands into PanDDA event maps (Pearce et al. 2017) using ligand coordinates and restraints generated with phenix.elbow (Moriarty et al. 2009).

Ligand Similarity, Chemical Diversity and Memorization Analyses

To test whether pose-prediction success reflected latent memorization, we quantified chemical similarity between test and training ligands using two complementary metrics. First, extended-connectivity fingerprints (ECFP4) were computed for all molecules, and Tanimoto coefficients (TC) were used to identify, for each test ligand, the most similar compound present in the training corpus (Chai-1 entries deposited before 1 December 2021 and AlphaFold3 entries before 30 September 2021). The training date cutoff for Boltz-2 is 1 June 2023 (STable 6).

As another measure of structural similarity, the maximum common substructure percentage (MCS%) was calculated using the rdFMCS algorithm in RDKit, reporting the proportion of heavy atoms shared between two ligands. To evaluate chemical diversity within benchmark datasets, pairwise TC or MCS% values were computed across all ligands. Best-First Clustering (BFC) based on these similarity metrics was used to estimate the number of unique chemotypes represented in each dataset. Principal component analysis (PCA) was used to visualize the resulting chemical space, and to identify an appropriate threshold for MCS-based scaffold grouping. Within the Mac1 benchmark set, applying an MCS cutoff of 35% yielded four distinct scaffolds, in agreement with PCA-derived cluster separation. The same clustering threshold was subsequently applied to define scaffold families for the AmpC, σ2 and D4 benchmark datasets.

Molecular Docking of Mac1 compounds

The modelled orthosteric site of the Mac1 model was available from the PanDDA analysis group deposition in the Protein Data Bank (PDB), with PDB ID: 5SQW (Gahbauer et al. 2023). We used this structure because the inhibitor (Z5014193706) was the most potent molecule with a structure determined around the same time as the ligands in this dataset were tested. This structure was used to generate matching spheres, which are later used by DOCK3.7 to pre-sample ligand conformations into binding sites by superimposing ligand atoms on spheres in the site. 45 matching spheres were used, and the docked ligands were scored by summing receptor-ligand electrostatic and van der Waals energies and corrected for context-dependent ligand desolvation. Receptor structures were protonated using Reduce. Partial charges from the united-atom AMBER force field were used for all receptor atoms. Potential energy grids for the different energy terms of the scoring function were precalculated using AMBER for the van der Waals term and the Poisson-Boltzmann method QNIFFT for electrostatics. Context-dependent ligand desolvation was calculated from the Generalized Born method. Ligands were protonated at pH7.4 (ChemAxon), and each protomer was rendered into 3D using Corina, followed by conformational sampling using Omega. For the docking campaign of 557 novel compounds, each molecule was sampled in about 2,719 orientations and, on average, 105 conformations by. Overall, over 290 million complexes were sampled and scored.

Co-folding

All three co-folding methods (AlphaFold3, Chai-1 and Boltz-2) were available from following sources:

AlphaFold3: https://github.com/google-deepmind/alphafold3

Chai-1: https://github.com/chaidiscovery/chai-lab.git

Boltz-2: https://github.com/jwohlwend/boltz

Ligand SMILES and amino acid sequences for the target protein were used to perform co-folding in an array of NVIDIA A40 GPUs. Multiple Sequence Alignment (MSA) was performed with the Jackhammer module. Co-folding tools will generate predicted protein-ligand complex poses (mmCIF files), and different confidence metrics for these predictions. Chai-1 and AlphaFold3 generated global confidence scores (pTM, ipTM), whilst AlphaFold3 had local atom-specific metrics, such as, pLDDT, PAE and contact probabilities in large vectors. To put an emphasis on prediction of ligands, we computed a ligand-centric pLDDT (L-pLDDT) by taking an arithmetic mean of pLDDT on ligand atoms, and a ligand-centric PAE (L-PAE) by taking an average of all PAE elements that contain ligand atoms. Minimum PAE (mPAE) was obtained by scanning all PAE elements that involve ligand atoms, and finding the smallest error. Boltz-2 generated all the confidence metrics in Chai-1 and AlphaFold3, but its affinity module generated an output of the binding likelihood (probability metric) and the predicted binding affinity (Boltz-2 pIC50) of compounds. Scripts to prepare input .json, .fasta files, post-process output files, are available in https://github.com/jongbin99/Cofolding.

Benchmarking Co-folding on Experimentally Validated Compounds

With novel compounds discovered by fragment-based screening, 557 Mac1 complexes were resolved by the crystallography work. The same sequences were used to perform co-folding with AlphaFold3, Chai-1 and Boltz-2. For further sets of analyses with docked hit lists on targets from the Large-Scale Docking (source: lsd.docking.org)(Hall et al. 2025), only AlphaFold3 and Boltz-2 were used. For each system, we redefined the threshold for true hits and non-binders as described herein:

AmpC β-lactamase: Among 1.7 billion docked molecules, the top scoring 1% were selected (n = 1,293) (Liu et al. 2025). We selected ones with apparent Ki < 400 μM (n = 247) as true hits, and the rest were regarded as non-binders.

σ2: True hits among these docked molecules were previously defined as displacing greater than 50% of 3H ditolylguanidine ([3H]DTG) to σ2 (Alon et al. 2021), but in our work, the cutoff is adjusted to 25% displacement, giving a less stringent threshold to include more molecules into the space of true hits. Single concentration point measurement (at 1 μM) is converted to an apparent Ki value using Cheng-Prusoff equation , with pre-calculated values of based on radioligand displacement (Alon et al. 2021). This yields a non-binding cutoff of an apparent Ki < 2.0 μM.

Dopamine D4: For the dopamine receptor D4, 138 million molecules were docked and then prioritized synthesizable high-ranking docked molecules (n = 549). True hits from these molecules were defined as displacing more than 50% of 3H-N-methylspiperone (Lyu et al. 2019), but similarly to σ2, we adjusted the cutoff to 25% displacement. Apparent Ki values are calculated similarly to σ2 case, but at 10 μM. This yields a non-binding cutoff of an apparent Ki < 7.8 μM.

Ligand Pose Evaluation via RMSD against crystal structures

Predicted ligand binding modes were benchmarked against experimentally determined reference poses from ultra-high-resolution (~1.0 Å) crystal structures. The Mac1 protein receptor was first superimposed by least-squares fitting of all Cα atoms using PyMol. The heavy-atom root-mean-square deviation (RMSD) between the aligned co-folded ligand and its crystallographic counterpart was then calculated. Before RMSD evaluation, both predicted and reference ligands were confirmed using a uniform RDKit-based sanitization pipeline (Buttenschoen et al. 2024). SMILES strings were converted to RDKit Mol objects, chemically validated (valence, aromaticity), and assigned bond orders from a canonical SMILES-derived template. We then check that ground-truth (experimental) and co-folded ligand coordinates for all molecules to have valid chemistry as described with the SMILES-derived template. A substructure match then enabled a one-to-one correspondence of heavy atoms between the ground-truth and co-folded poses for comparison. The crystal ligand and co-folded ligand had the center-of-mass (COM) calculated separately, and measured the distance between each COM. If the distance exceeds 2.5 Å, we would expect the RMSD to be large, indicating that ligand poses are not modelled accurately in an orthosteric pocket (Nittinger et al. 2025). For all the predicted ligand poses, rdMolAlign.CalcRMS from Rdkit.Chem is used for calculating the ligand RMSD between co-folding and the ground-truth crystal pose.

Hit-rate curves for different scoring methods

To get hit-rate curves for co-folding and docking scores across three targets of docked hit-lists (AmpC, σ2, D4), molecules in the list are ordered by increasing score (for DOCK) and decreasing score (for L-pLDDT and Boltz-2 pIC50). A rolling window of 100 was used for AmpC and σ2, and 50 for D4 (Liu et al. 2025), calculating the hit rate as a fraction of molecules with experimental binding affinity that match our definition for true hits described above (refer to Benchmarking Co-Folding on Experimentally Validated Compounds in Method section). The docked hit-list is a mixture of known actives and non-binders that include false positives based on docking scores. For direct comparison of hit classification performance between DOCK energies, AF3 L-pLDDT and Boltz-2 pIC50, the pProp (the negative base 10 logarithm of the fractional rank) was assigned within the list, instead of the billion-scale full docking list. For each window, we used a 95% Wilson confidence interval with normal approximation for the hit rate.

Statistical analysis

Pearson’s correlation coefficients were used throughout the paper to quantify linear associations between two variables in the study. We have also reported a mean absolute error of data points with respect to the regression line for Boltz-2 affinity pIC50 against the measured pIC50. The mean absolute errors are also calculated for the perfect fit (Pearson r = 1) and for the fixed measured pIC50 (Pearson r = 0), to give an indication of how accurate the Boltz-2 scores are with respect to actual binding energies (in kcal), in terms of correlations and absolute errors. Friedman and post-hoc Conover tests were used to compare how each of AF3, Boltz-2 and DOCK3.7 score predicts for actual binding affinity, and Holm-Bonferroni correction for pairwise comparison p-values (Ash et al. 2025). The distribution of scores for known actives and docked false positives were compared using the Kolmogorov-Smirnov (KS) non-parametric statistical test, as shown in violin plots, and the KS statistic value indicates how distant the two distributions are, and p-values show the extent of statistically detectable differences. All analyses were performed using Python packages.

eLife Assessment

This study addresses an important gap in drug discovery by delivering a rigorous, large-scale evaluation of widely used co-folding methods for predicting ligand-bound protein complexes and virtual screening. A key strength is the comprehensive benchmarking framework, which leverages structures and chemical compounds that were absent from the AI models training set, thereby providing particularly compelling and unbiased evidence of co-folding performance. The findings clearly delineate the complementary roles of deep learning-based co-folding and physics-based docking, offering practical guidance for their rational integration into drug discovery workflows. Although the conclusions are convincing, improvements in the test cases, presentation, and usability can further strengthen the overall impact.

Reviewer #1 (Public review):

The authors conducted a comprehensive benchmarking and evaluation of co-folding platforms, including AlphaFold3, Boltz-2, Chai-1, and the docking algorithm Dock3.7, which employs a physics-based scoring function that incorporates van der Waals interactions, electrostatics, and ligand desolvation energies. The system of interest was the SARS-CoV-2 NSP3 macrodomain (Mac1), an increasingly popular antiviral target, and the ligand sets comprised 557 unseen ligand poses (keeping the training for these co-folding platforms in mind). Additionally, the authors investigated whether the co-folding models could distinguish true ligands from non-binding small molecules. The study is thorough, with extensive statistical support and consensus across multiple metrics (chemoinformatics for quantifying ligand similarity and efficacy). The questions that the authors aim to address are whether the co-folding models struggle with memorization, whether they can distinguish between a true and a false binder, whether they replicate experimental binding affinities and efficacy, and how they compare to the physics-based docking algorithm (Dock3.7).

Strengths:

Overall, this is a scientifically solid paper. The work is highly detailed and well executed, featuring thorough data analysis and statistical assessment.

Weaknesses:

My main concern is that the study's aim is a bit unclear. Modern benchmarking studies comparing physics-based docking with deep learning-based co-folding approaches (e.g., AF3, Boltz-2, Chai-1, and others) are increasingly expected to go beyond aggregate performance metrics. In addition to rigorous dataset construction, transparent methodology, and appropriate statistical evaluation, high-impact benchmarks typically provide actionable guidance on when each method class is most appropriate, reflecting their distinct inductive biases and practical constraints. Failure-mode analyses that link performance differences to protein flexibility, ligand chemistry, or binding-site characteristics are particularly valuable, as they move comparisons beyond "scoreboard" assessments toward mechanistic understanding. While full biological validation is not expected, qualitative interpretation grounded in physical and biological principles strengthens conclusions. Providing reproducible workflows or reference pipelines is not mandatory, but it is increasingly viewed as a best practice because it facilitates adoption and helps contextualize results for practitioners.

Reviewer #2 (Public review):

Summary:

The manuscript by Kim et al. evaluates the performance of three modern AI-based methods in predicting complex structures and binding affinities between proteins and chemical compounds. An honest 'prospective' evaluation is achieved by studying benchmark structures and chemical compounds that did not exist in the PDB at the time the AI structure prediction models (AlphaFold3, Chai-1, Boltz-2) were trained.

Strengths:

(1) The study addresses an important question in modern computational biology and drug discovery, and establishes the strengths and limitations of the three tools in solving various computational chemistry tasks, including compound pose prediction, active-inactive discrimination, and potency ranking.

(2) The conclusions are based on examination of four separate targets and respective compound datasets, where for one of the targets, the authors also obtained numerous X-ray structures to serve as experimental answers for the binding pose prediction task.

(3) The study reports relationships between structure prediction confidence, predicted energies (DOCK3.7), and affinity predictions (Boltz-2) with the geometric accuracy of compound pose prediction as well as the experimentally measured potency.

(4) One of the key findings is the limited ability of co-folding methods to predict conformational rearrangements, which does not correlate with their ability to predict binding poses of the compounds inducing these rearrangements.

(5) The findings could serve as useful guidelines for computational chemists in selecting appropriate software and scoring schemes for each task.

Weaknesses:

While I consider this a solid study, several aspects would need to be addressed to make it really strong:

(1) DOCK3.7 docking and scoring experiments were performed using one experimental structure of Mac1, selected from dozens of structures based on a criterion that is not sufficiently well justified. For sigma2 receptor, dopamine D4 receptor, and AmpC β-lactamase, it is not clear which structures or models were selected for docking at all. It is well known that geometry predictions, scoring, and active-inactive ROC AUCs are all strongly influenced by the selected structure. It would be important to attempt Mac1 docking using all available experimental Mac1 structures, or at least against representative structures in various conformations; it would also be quite insightful to compare results to docking of the same compound sets to AF3, Boltz-2 and Chai-1 predicted structures of Mac1. Same goes for the docking studies of sigma2, D4, and AmpC β-lactamase.

(2) For binding affinity predictions, as a control, authors should consider compound co-folding with an unrelated protein, or even with a pseudo-peptide that consists of a few random single amino acids - this would provide an honest baseline for such predictions.

(3) ROC curves Figure 3 and elsewhere should be shown, and AUCs quantified/reported on a log or square-root scaled x-axis, to emphasize early enrichment, which is the area of practical significance for these predictions. For example, Figure 3A currently suggests that the pose prediction performance of AF3 exceeds that of Boltz-2 whereas the early enrichment is clearly better for Boltz-2.

(4) 'Trained set' in figures and text should probably be 'training set'? Or otherwise explain this new term the first time it is introduced.

(5) Figure 1 illustrates a projection onto the first two principal components of a space that apparently had only one (scalar) metric for each compound pair (% maximum common substructure or Tanimoto coefficient); the authors need to better explain the principle behind this analysis and visualization.

Reviewer #3 (Public review):

Summary:

This study's core conclusions are well-supported by data. It is shown that co-folding outperforms docking in known ligand pose/affinity prediction (validated by RMSD and IC₅₀ correlation), struggles with false-positive discrimination in virtual screens (lower AUC values), and is complementary to docking (non-correlated errors, distinct strengths in drug discovery stages).

Strengths:

(1) Unprecedented prospective design with 557 novel Mac1-ligand complexes ensures rigorous, independent evaluation of co-folding methods.

(2) Comprehensive comparison of 3 co-folding tools (AlphaFold3, Chai-1, Boltz-2) with DOCK3.7 across diverse targets and metrics enables nuanced performance assessment.

(3) The study clearly demonstrates complementary roles of co-folding (superior pose/affinity prediction for known ligands) and docking (better hit prioritization), and addresses deep learning memorization concerns via ligand similarity analysis.

Weaknesses:

(1) Limited generalization to diverse protein families (e.g., no ion channels/transporters).

(2) Ambiguity in the mechanism underlying co-folding's failure to predict rare conformational changes.

(3) Virtual screen comparison is unbalanced (docking-prioritized hit lists bias results).

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

The authors conducted a comprehensive benchmarking and evaluation of co-folding platforms, including AlphaFold3, Boltz-2, Chai-1, and the docking algorithm Dock3.7, which employs a physics-based scoring function that incorporates van der Waals interactions, electrostatics, and ligand desolvation energies. The system of interest was the SARS-CoV-2 NSP3 macrodomain (Mac1), an increasingly popular antiviral target, and the ligand sets comprised 557 unseen ligand poses (keeping the training for these co-folding platforms in mind). Additionally, the authors investigated whether the co-folding models could distinguish true ligands from non-binding small molecules. The study is thorough, with extensive statistical support and consensus across multiple metrics (chemoinformatics for quantifying ligand similarity and efficacy). The questions that the authors aim to address are whether the co-folding models struggle with memorization, whether they can distinguish between a true and a false binder, whether they replicate experimental binding affinities and efficacy, and how they compare to the physics-based docking algorithm (Dock3.7).

We thank Reviewer 1 for this thoughtful summary of our work.

Strengths:

Overall, this is a scientifically solid paper. The work is highly detailed and well executed, featuring thorough data analysis and statistical assessment.

Weaknesses:

My main concern is that the study's aim is a bit unclear. Modern benchmarking studies comparing physics-based docking with deep learning-based co-folding approaches (e.g., AF3, Boltz-2, Chai-1, and others) are increasingly expected to go beyond aggregate performance metrics.

Indeed, we have gone into several examples of failures and successes for each of these methods. As we are not developing these methods ourselves, we also think this dataset will be a valuable contribution for improving them further.

In addition to rigorous dataset construction, transparent methodology, and appropriate statistical evaluation, high-impact benchmarks typically provide actionable guidance on when each method class is most appropriate, reflecting their distinct inductive biases and practical constraints. Failure-mode analyses that link performance differences to protein flexibility, ligand chemistry, or binding-site characteristics are particularly valuable, as they move comparisons beyond "scoreboard" assessments toward mechanistic understanding.

Right now, we do not observe meaningful trends that separate the failure modes for any individual method. This is covered in Supplementary Figures 6 and 7.

While full biological validation is not expected, qualitative interpretation grounded in physical and biological principles strengthens conclusions. Providing reproducible workflows or reference pipelines is not mandatory, but it is increasingly viewed as a best practice because it facilitates adoption and helps contextualize results for practitioners.

We note that our code is available (https://github.com/jongbin99/Cofolding/) and all structural data will be publicly accessible in the PDB alongside publication (we only held it back only for “blinding” during peer review to avoid contamination with any new deep learning methods).

Reviewer #2 (Public review):

Summary:

The manuscript by Kim et al. evaluates the performance of three modern AI-based methods in predicting complex structures and binding affinities between proteins and chemical compounds. An honest 'prospective' evaluation is achieved by studying benchmark structures and chemical compounds that did not exist in the PDB at the time the AI structure prediction models (AlphaFold3, Chai-1, Boltz-2) were trained.

Strengths:

(1) The study addresses an important question in modern computational biology and drug discovery, and establishes the strengths and limitations of the three tools in solving various computational chemistry tasks, including compound pose prediction, active-inactive discrimination, and potency ranking.

(2) The conclusions are based on examination of four separate targets and respective compound datasets, where for one of the targets, the authors also obtained numerous X-ray structures to serve as experimental answers for the binding pose prediction task.

(3) The study reports relationships between structure prediction confidence, predicted energies (DOCK3.7), and affinity predictions (Boltz-2) with the geometric accuracy of compound pose prediction as well as the experimentally measured potency.

(4) One of the key findings is the limited ability of co-folding methods to predict conformational rearrangements, which does not correlate with their ability to predict binding poses of the compounds inducing these rearrangements.

(5) The findings could serve as useful guidelines for computational chemists in selecting appropriate software and scoring schemes for each task.

We appreciate Reviewer 2’s summary of the novelty of the dataset and analysis.

Weaknesses:

While I consider this a solid study, several aspects would need to be addressed to make it really strong:

(1) DOCK3.7 docking and scoring experiments were performed using one experimental structure of Mac1, selected from dozens of structures based on a criterion that is not sufficiently well justified. For sigma2 receptor, dopamine D4 receptor, and AmpC β-lactamase, it is not clear which structures or models were selected for docking at all. It is well known that geometry predictions, scoring, and active-inactive ROC AUCs are all strongly influenced by the selected structure. It would be important to attempt Mac1 docking using all available experimental Mac1 structures, or at least against representative structures in various conformations; it would also be quite insightful to compare results to docking of the same compound sets to AF3, Boltz-2 and Chai-1 predicted structures of Mac1. Same goes for the docking studies of sigma2, D4, and AmpC β-lactamase.

In any program, a decision has to be made as to which template will be used for docking, we justified the choice in the methods:

“We used this structure because the inhibitor (Z5014193706) was the most potent molecule with a structure determined around the same time as the ligands in this dataset were tested.”

We stand by this as a reasonable assumption. Similarly, for sigma2, D4, and AmpC β-lactamase, the template was chosen in the respective papers:

a) The σ2 receptor bound to cholesterol (PDB ID: 7MFI) was used in the docking calculations.

- This structure was determined in the paper, the first structure of sigma2 and therefore a worthy template

b) The D4 receptor campaign used PDB 5WIU

- This was one of two D4 structures available and chosen because it was not bound to sodium

c) For AmpC, the campaign used the structure in the Protein Data Bank (PDB) 1L2S

- This maximizes comparisons to other docking studies that used the same receptor template.

The major goal of this study is to compare different methods under reasonable (but perhaps as the reviewer points out, not optimal) conditions, not to optimize docking score.

(2) For binding affinity predictions, as a control, authors should consider compound co-folding with an unrelated protein, or even with a pseudo-peptide that consists of a few random single amino acids - this would provide an honest baseline for such predictions.

This suggestion would be valuable for understanding the performance for these methods from the perspective of ligand specificity (a valuable, but separate, goal). Surely this will generate some number or some prediction - but what would this baseline mean and how would it be relevant for drug discovery? Therefore, we do not think this suggestion is relevant for the issues being investigated in this manuscript.

(3) ROC curves Figure 3 and elsewhere should be shown, and AUCs quantified/reported on a log or square-root scaled x-axis, to emphasize early enrichment, which is the area of practical significance for these predictions. For example, Figure 3A currently suggests that the pose prediction performance of AF3 exceeds that of Boltz-2 whereas the early enrichment is clearly better for Boltz-2.

We agree with this, and added a semi-logAUC plot for Figure 3A. For Figure 5, we also generated a semi-logAUC plot to see early ligand enrichment clearly, added as Supplementary Figure 11. We added the text:

“Considering its early enrichment performance, Boltz-2 Ligand ipTM was the strongest predictor of pose accuracy based on normalized logAUC (20.5% above random, Fig. 3a). In contrast, although Boltz-2 pIC50 showed poor overall discrimination, it overestimated its ability to enrich true positive poses at low false positive rates, despite having a weak early enrichment behavior”

(4) 'Trained set' in figures and text should probably be 'training set'? Or otherwise explain this new term the first time it is introduced.

Thank you for pointing out this for clarification. ‘Training set’ is the correct word, and we made changes appropriately across all figures and texts.

(5) Figure 1 illustrates a projection onto the first two principal components of a space that apparently had only one (scalar) metric for each compound pair (% maximum common substructure or Tanimoto coefficient); the authors need to better explain the principle behind this analysis and visualization.

This suggestion is valuable, since we often use PCA to reduce dimensionality for more complex features. For clarification, we actually have a full pairwise similarity matrix for all tested Mac1 compounds based on each of Tc and MCS%. PCA for each MCS% and Tc is a representation of each pairwise similarity matrix. We also made a change in Figure 1 caption to make this point clearer:

“projection of compounds represented by their full pairwise similarity vectors (by ECFP-4 Tc and MCS%)”

Reviewer #3 (Public review):

Summary:

This study's core conclusions are well-supported by data. It is shown that co-folding outperforms docking in known ligand pose/affinity prediction (validated by RMSD and IC₅₀ correlation), struggles with false-positive discrimination in virtual screens (lower AUC values), and is complementary to docking (non-correlated errors, distinct strengths in drug discovery stages).

Strengths:

(1) Unprecedented prospective design with 557 novel Mac1-ligand complexes ensures rigorous, independent evaluation of co-folding methods.

(2) Comprehensive comparison of 3 co-folding tools (AlphaFold3, Chai-1, Boltz-2) with DOCK3.7 across diverse targets and metrics enables nuanced performance assessment.

(3) The study clearly demonstrates complementary roles of co-folding (superior pose/affinity prediction for known ligands) and docking (better hit prioritization), and addresses deep learning memorization concerns via ligand similarity analysis.

We thank Reviewer 3 for pointing out the unprecedented and comprehensive nature of our study

Weaknesses:

(1) Limited generalization to diverse protein families (e.g., no ion channels/transporters).

We agree - we have not explored the entire proteome and these are important target classes that will surely be investigated by future studies. We focused on targets here where we had large number of X-ray crystal structures (Mac1) and affinity/inhibition measurements from docking (the other three targets).

(2) Ambiguity in the mechanism underlying co-folding's failure to predict rare conformational changes.

Again, we agree. We are not the developers of these methods. We observe that these methods do not predict conformational changes with high fidelity and this weakness is an area that co-folding methods will surely prioritize in the future.

(3) Virtual screen comparison is unbalanced (docking-prioritized hit lists bias results).

We acknowledge this in the results: “An important caveat is that the hit-lists were composed of molecules prioritized by docking in the first place, giving it an advantage on these particular sets.” and discussion: “Finally, comparing co-folding to docking based on hit-lists themselves selected by docking is arguably unfair to co-folding. Counter-balancing this is the inclusion, in each of the three hit lists, of molecules that had mediocre and poor docking scores intentionally selected to test the correlation between docking score and hit-rate. Here too, the correlation between co-folding score and likelihood to bind, what we sometimes call a “dock-response-curve” was no better than docking’s, often worse (SFig.11).”

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Here are suggestions for revisions:

(1) The writing is at times obtuse and hard to follow.

This happens sometimes when multiple authors are writing together. We apologize and are happy to respond to specific areas that can be streamlined to be easier to follow.

(2) In the Results section, "A set of 557 previously unreported Mac1 ligand complexes", the authors have compared the ligand poses across different metrics such as Tc - a standard, highly effective method in chemo-informatics and MCS (maximum common substructures); these are standard metrics for quantifying the structural similarity between pairs of small molecules. This part of the analysis checks whether this is memorization; it is critical to compare the two metrics, but it is not sufficient to draw a conclusion.

Thank you for pointing out about the structural similarity of molecules co-folded to those present in the training set (resolved as Mac1 complexes and deposited in PDB before training dates). We have conducted an analysis where we do a pairwise similarity comparison for all ligands present in the PDB (regardless of the target), by both Tc and MCS, and overlay the cluster of ligands we tested (Mac1, AmpC, sigma2, D4). This should show where our tested benchmark datasets lie in the chemical space covered in the entire PDB. Each cluster (around 500 to 1300 compounds per target system) is overlaid on the cluster of all ligands deposited in PDB (over 50,000 compounds), and each cluster was relatively diverse by both Tc and MCS.

(3) In the "Co folding can accurately reproduce poses of ligands dissimilar to those trained." Subsection under Results, the authors' conclusions are hard to follow; they state that the co-folding models often mispredict or miss the alternative conformation, but they also predict poses that are distinct from the training set. What does that imply?

Our interpretation is actually a somewhat unsettling one: co-folding gets the ligand pose right even when it gets the protein wrong, and even when the ligand is novel. This suggests the models may be anchoring on conserved pharmacophoric interactions (like the adenosine-mimicking purine scaffold) rather than truly modeling the physics of the full complex. We added to the results section:

This result suggests that co-folding reliably recapitulates dominant ligand-binding interactions even in the absence of accurate protein conformational modeling, providing further support to the idea that they are learning specific interaction patterns rather than a deeper physics-based representation (Masters et al. 2025).

(4) The Discussion section connects the results and conclusions, but it can be challenging to grasp the study's overall message.

We think the final paragraph hits on three major points:

- Co-folding accurately predicts ligand poses for known binders, but fails to capture conformational changes

- Co-folding does not reliably distinguish true binders from false positives in virtual screening hit lists

- Docking and co-folding are complementary rather than competing tools

(5) The work is highly detailed and well executed, featuring thorough data analysis and statistical assessment. The value of the paper would be further enhanced by explaining how it differs from seemingly similar results reported in other studies, including the one cited in this manuscript (see https://www.biorxiv.org/content/10.64898/2025.12.04.692352v1).

The Mac1 results are completely unique. However, the docking datasets are exactly the same as those analyzed in the Menon et al manuscript. We don’t think our results differs from conclusions of the Menon et al manuscript as we wrote: These observations are supported by a fascinating study on some of the same ligand sets as investigated here, using AlphaFold3, reaching similar conclusions (Menon et al. 2025).

Reviewer #3 (Recommendations for the authors):

(1) Expand target diversity to include ion channels, transporters, etc., beyond enzymes and GPCRs.

(2) Investigate the cause of co-folding's failure in predicting rare conformational changes (e.g., adjust sampling, MSA inputs, or add experimental constraints).

(3) Mitigate docking bias in virtual screens (e.g., re-analyze unbiased compound libraries).

We addressed these three points in the public review above

(4) Test Boltz-2's affinity predictions without linear calibration and compare with FEP.

The data without linear calibration are included in the manuscript. Comparing such a large number of compounds with FEP is currently beyond our capabilities.

(5) Conduct proof-of-concept to test co-folding-docking integration for better hit rates.

We think this is well beyond the scope of this manuscript - but look forward to testing this idea in the future.

We also got one community review that we respond to below:

Summary

This manuscript evaluates the performance of co-folding models when tasked with 1) the recapitulation of a large number of experimentally determined co-crystal structures of Mac1 with a series of Mac1 ligands and 2) the rescoring of hits to identify false positives originally derived from a set of large docking-based virtual screens. The evaluation leverages a dataset of crystal structures and affinity data from high-throughput crystallographic and biophysical screens, respectively. These data uniquely enable this report to focus on the ability of co-folding models to handle ligands, resulting in an analysis that is particularly timely given the wide adoption of co-folding models and the relative scarcity of such ligand-focused benchmarks among existing evaluations, which have primarily focused on protein structure prediction or binder design.

Thank you for this thoughtful summary of our work

Feedback

The experiments and analyses in the manuscript are well thought-out and do not have any significant issues. There are a few high-level points that may improve the clarity and completeness of the results. Importantly, none of the suggested additional experiments will affect the conclusions of the paper, but rather help provide additional context for the results:

The first section presents an exciting opportunity to frame the Mac1 ligands against ligands in the PDB more broadly. It would be informative to assess whether chemotypes that are easier or harder to predict accurately and confidently are over- or under-represented in the PDB as a whole. Note that this is not a recommendation that new scaffold similarity metrics be incorporated into the analysis, but rather that analyses similar to those already performed in the manuscript are performed using all ligands in the PDB. For example, PCA-based analyses similar to those in Fig. 1c could be used to examine Mac1 ligands in the context of all PDB ligands enabling questions such as whether similarity to a nearest PDB neighbor, cluster size in a Tc/MCS PCA space, or other frequency-based measures show any relationship with prediction vs. crystal structure RMSD. Such analyses could provide additional insight into how effectively models leverage ligand information present in the PDB overall, as opposed to biases arising specifically from scaffolds represented in Mac1 structures in the PDB, which are already well covered in the manuscript. The conclusion that Tc/MCS do not correlate with the ligand RMSDs for the ligands already associated with the Mac1 is well supported, and presumably suggests that a correlation would not exist against the backdrop of the PDB, but it would be interesting to see the data using analyses similar to those already done in the manuscript nonetheless.

We are adding new figures in SFig.1 that consider how different clusters of ligands tested for our co-folding analysis are distributed across the chemical space in PDB. This is done by making a similarity comparison between every ligand in PDB and those tested in our analysis by Tc and MCS%, then plotting in PCA space for each metric. We are excited to see that each dataset covers a wide scope in PCA space, but at the same time, there are unexplored areas in the chemical space of PDB by co-folding.

Similarly, even though the four proteins used in this manuscript are not themselves the primary focus of the analysis, it would be valuable to perform a high-level assessment of the precedent for each protein in the PDB (beyond the count of liganded structures in Table S6), either in protein sequence space (e.g., MSAs) or structural space (e.g., FoldSeek). An analysis like this would provide important context about whether any of the proteins in the study have close homologs with liganded structures in the PDB, or are generally overrepresented in the PDB. The fact that the AUC for L-pLDDT for AmpC is higher than σ2 and D4, for example, is notable given the relative abundance of liganded AmpC structures in the PDB (this raises potentially interesting questions related to where DOCK3.7 and AF3 actually place the ligands, given the orthosteric β-lactam binding pocket in AmpC, although this is outside of the scope of this manuscript).

High-level assessment of the precedent for each protein in the PDB will definitely help to understand if proteins we used have close homologs with liganded structures in the PDB. Our Supplementary Table 6 covers the extent to which these liganded structures were available by cutoff dates for AF3, Chai-1 and Boltz-2. AmpC had more homologs than sigma2 and D4, and this may explain a better AUC for AF3 L-pLDDT specifically for this target.

A discussion of the affinity probability results (`affinity_probability_binary`) from Boltz-2 is likely warranted in the second section in addition to the pIC50s that are already reported (`affinity_pred_value`). The former seems like it would be more applicable for section 2 of the manuscript, but both warrant inclusion—they should both be calculated by default when the affinity pipeline in Boltz-2 is turned on, so it wouldn't involve any more inference.

As boltz-2 affinity module outputs both affinity probability binary output and affinity predicted value, we kept track of both metrics. So we tried re-ranking hit lists using both metrics. Where boltz-2 performed better (Sigma2, D4), binary probability values were more representative as a metric to differentiate true actives from non-binders. This was more clear in semi-logarithmic ROC plots. However, in AmpC, both Boltz-2 scoring metrics performed similarly. Such inconsistency in trend made it difficult to draw conclusions.

Minor points

A more detailed description of the experimental methods used to generate the ground-truth data in the introduction (even though these have been explained in prior works) would help orient the reader early on, and ground the benchmarking aspect of the story. In general, the abstract and introduction would benefit from a more cohesive through-line to tie the two complementary but orthogonal sections of the paper together.

We will include a more thorough description alongside the PDB depositions. As for the two sections, we have tried to tie them together from the perspective of drug discovery workflows…

The cutoffs in the "Co-folding can accurately reproduce..." section shift between 2.5 Å (from the ligand center of mass) and 2.0 Å. Is there a reason for this? Along similar lines, mentioning cutoffs for true positives/negatives when introducing the ROC analyses later on in the Mac1 section seems unnecessary since no cutoff should be necessary here.

We used 2.5A distance to COM to just get at “broadly the correct binding site” for fast filtering and 2.0A RMSD because that is the broadly accepted standard in the field for “relatively correct binding pose”.

Version 2

Abstract

Accurate prediction of ligand-bound protein complexes and ranking them by affinity are central problems in drug discovery. While deep learning co-folding methods can help address these challenges, their evaluation has been hampered by the difficulties in assessing independence from training data and insufficiently large test sets. Here we test the ability of co-folding methods to predict the structures of 551 ligands, 489 of which are of sufficient quality for evaluation, bound to the SARS-CoV-2 NSP3 macrodomain (Mac1) that were determined after the training cut-off dates. AlphaFold3 (AF3), Boltz-2, and Chai-1 each reproduced >50% of the Mac1 ligand poses to better than 2 Å RMSD of experiment. Despite the potential for co-folding to describe protein conformational changes that stabilize ligand binding, we did not find that common conformational rearrangements, including peptide flip and a large loop opening, were recapitulated by the co-folding prediction. For AF3 and Chai-1, ligand pose prediction confidence weakly, but significantly, tracked experimental potency, while DOCK3.7 energies were only weakly correlated. Boltz-2 affinity predictions showed the strongest correlation with measured potency and, after calibration, achieved lower mean absolute error than a baseline predictor. We next assessed whether co-folding scores could rescore docking hit-lists to distinguish true ligands from non-binders among hundreds of molecules prospectively experimentally tested against AmpC β-lactamase, the dopamine D4 and the σ2 receptors. AF3 ligand pose confidence values did not separate true ligands from high-scoring false-positives as effectively as docking scores or Boltz-2 affinity predictions did. Taken together, the modest, but independent correlations of docking score and co-folding confidence or affinity suggests that integrating physics-based and deep-learning and approaches may help with hit prioritization and subsequent optimization in structure-based ligand discovery.

Introduction

Deep Learning methods such as AlphaFold2 (Jumper et al. 2021) and RosettaFold (Baek et al. 2021) revolutionized protein structure prediction. Key to their wide adoption has been their extensive testing, first in well-controlled prospective tests like CASP (Kryshtafovych et al. 2021; Ashizawa et al. 2025) and subsequently in many studies where the methods found new applications and new experimental validation (Scardino et al. 2023; Goverde et al. 2023). Notably, these methods can provide protein structural models for small molecule ligand discovery by docking/virtual screening, leveraging physics-inspired functions to score the resulting complexes (Seok and Tiwary 2025; Brocidiacono et al. 2025; Huang et al. 2006; Michino et al. 2025; Xie et al. 2025). Retrospective analyses of AlphaFold2 models suggested that the accuracy of the predicted ligand binding poses was lower than experimental structures (Karelina et al. 2023; Holcomb et al. 2023). However, prospective use of AlphaFold2 models and experimental structures (determined after the training cutoff) to identify new ligands revealed that both approaches yielded similar quality and quantity of ligands (Lyu et al. 2024; Díaz-Holguín et al. 2024). These results indicate that AF2 models may sample alternative low energy conformations that are compatible with distinct ligands from the experimentally resolved conformations, reconciling the poor retrospective recall with the prospective successes.

More recently, the advent of co-folding methods like AlphaFold3 (AF3) (Abramson et al. 2024), Chai (Boitreaud et al. 2024), and Boltz (Wohlwend et al. 2025) that leverage diffusion-based approaches for multi-component (e.g. protein and small molecule) complex predictions, have promised to extend the impact of deep learning methods from protein structure prediction to small molecule ligand-bound protein complexes (He et al. 2025). The advantages co-folding may have over traditional docking include its abilities to explore conformational variability of the protein on-the-fly and to predict ligand affinity, reportedly as well as or better than a gold-standard physics-based method, free energy perturbation (FEP) (Passaro et al. 2025). However, multiple studies have highlighted how co-folding success can track with structural similarity to training structures, consistent with substantial structural memorization (Xu et al. 2025; Škrinjar et al. 2025). An innovative approach introduced adversarial mutations to fill binding sites to show that co-folding models will still produce high confidence poses even when interactions are physically impossible (Masters et al. 2025). These results highlight pervasive memorization and hallucination in current co-folding methods and suggest that alternative computational approaches are needed to quantify the physical basis of the protein-ligand interactions.

Still, co-folding has potential for synergizing with many aspects of structure-based design and more traditional computational chemistry approaches. Two attractive applications are the prediction of ligand-bound complexes, around which structure-based ligand design is articulated (Erlanson et al. 2025); (Backus et al. 2016), and the re-scoring of large library docking hitlists. Co-folding is orders of magnitude slower than most docking programs, which can address libraries of hundreds-of-millions to billions of compounds on CPU clusters. The method nevertheless is fast enough to re-rank virtual screening “hit-lists” and may be better at categorizing certain types of ligands (e.g. those that stabilize alternative conformations) than the simpler physics-based methods. Moreover, the errors and biases from co-folding affinity estimation methods (Passaro et al. 2025) may be independent from those in docking, suggesting that each method may have different strengths in distinguishing likely from unlikely ligands. However, partly because of the lack of widely adopted community prospective tests, akin to CASP (Gathiaka et al. 2016) and partly because of the challenges of developing, testing, and potentially optimizing new chemical matter for ligand activity studies (Ackloo et al. 2022), co-folding methods have not yet been subject to the rigorous, large-scale testing that helped establish the utility, domain of applicability, and ultimately optimization of the early protein structure-prediction methods.

It therefore seemed interesting to test co-folding predictions of the co-complexed structures and affinities of ligands in a drug-discovery series from outside of their training sets, and to test their ability to distinguish likely from unlikely ligands from large library virtual screens (Cabeza de Vaca et al. 2025). Here, we evaluated the performance of three widely-used co-folding methods, AF3, Chai-1, and Boltz-2 against two benchmarks that speak to domain utility and are largely outside the benchmarks against which they trained. First, we tested the ability of the co-folding methods to recapitulate the crystallographic structures of 551 different ligands bound to the SARS-CoV-2 Macrodomain (Mac1), an emerging antiviral target (Suryawanshi et al. 2025). This dataset represents the scale and depth typical of the early phase of industrial structure-based drug design campaigns. Second, we investigated the ability of the co-folding methods to distinguish true-from false-positives in hit lists from recent large library, prospective docking against three well-behaved targets: AmpC β-lactamase, the σ2 receptor, and the dopamine D4 receptor (Lyu et al. 2019; Alon et al. 2021; Liu et al. 2025). In the campaigns against these targets, between 500 and 1500 molecules were prospectively tested across a range of docking ranks identifying ligands with previously unexplored chemotypes as well as finding many docking false positives. Contemporaneous work has highlighted the importance of using AF3 in the same experimental datasets from ultra-large library docking campaigns (Menon et al. 2025). Our analyses reveal the strengths and limitations of different co-folding methods versus traditional docking and point to potential synergies that can improve probe and lead discovery pipelines.

Results

A set of 551 previously unreported Mac1 ligands complexes

As part of a drug discovery effort (Suryawanshi et al. 2025; Jaishankar et al. 2025; Correy et al. 2025; Flowers et al. 2025; Gahbauer et al. 2023; Schuller et al. 2021), we tested thousands of novel small molecules as inhibitors of Mac1, a new target for SARS-CoV2 and other viruses. This led to potent inhibitors that are active in vivo as anti-virals (Suryawanshi et al. 2025; Jaishankar et al. 2025). As part of this work, we determined x-ray crystal structures for ∼1000 small molecules bound to Mac1, typically at close to 1 Å resolution, but only a limited number of these were disclosed in the papers describing the more advanced candidates. Moreover, only up to 326 of our molecules, largely consisting of initial fragments and early hit-to-lead material, were available in public databases prior to the training cutoffs for the cofolding methods (e.g. 2021-09-30 for AF3, 2023-06-30 for Boltz-2). Apart from their role in driving inhibitor optimization for Mac1, we thought that these structures might be a resource for evaluating computational approaches against a set that is representative of the scale and depth often seen in industrial structure-based drug design campaigns.

From our internal dataset, we identified 551 ligands where we determined X-ray structures that have not been disclosed previously (Fig. 1a, Supplementary File 1). These molecules all bind in the active site where they compete with the substrate, ADP-ribosylated (ADP-r) peptides. There are several stereotyped interactions across this set, which have been described in previous fragment and lead optimization efforts (Gahbauer et al. 2023; Schuller et al. 2021; Correy et al. 2025), including: interactions with backbone NH groups of Phe156 and Asp157 (the oxyanion site), aromatic packing against the side chain of Phe156 (mimicking the adenosine ring of ADP-r), and hydrogen bonds to Asp22 and Ile23 (the upper active site) (Fig. 1a). Because these ligands were synthesized during an active campaign and many did not meet evolving thresholds for single point % inhibition, only 202 have fully measured dose response IC50 estimations by a HTRF-based peptide displacement assay. For these compounds, the measured potencies range from sub-micromolar to >500 uM (Fig. 1b). This distribution is skewed towards weaker affinities as more potent molecules were progressed and disclosed in papers describing in vivo-active lead molecules (Suryawanshi et al. 2025; Jaishankar et al. 2025).

Structure diversity of 551 previously unreported ligands co-folded with the SARS-CoV-2 Mac1 (NSP3 macrodomain 1).

a) Global superposition of all 551 ligands within the Mac1 binding pocket highlighting residues involved in key interactions. b) The affinity distribution of 202 molecules determined by HTRF (drug-response curves at 8 concentration points). c) Principal Component Analysis projection of compounds represented by their full pairwise similarity vectors (by ECFP-4 Tc and MCS%). d) The highest MCS% vs. ECFP-4 Tc values in the training set for each Mac1 compound. At right are examples of the low similarity, scaffold hop, and high similarity distributions, respectively.

The diversity of the 551 ligands is important for understanding the performance of the cofolding methods. By topological similarity, using ECFP4-based pairwise Tanimoto coefficients (Tc), the ligands were diverse, with pairwise Tc clustering between 0.2 to 0.4 (Figure 1—figure supplement 1a, Supplementary File 2). For context, a scaffold-hop for the ECFP4-based fingerprint is thought to occur at Tc values below 0.35 (Cramer et al. 2004; Muchmore et al. 2008), while random similarity occurs around Tc values of 0.25 (Hert et al. 2008). As the Tc metric has well-known liabilities that can mask what a chemist would see as similarity, we also calculated pairwise Maximum Common Substructures (MCS) and found a distribution that centers around 44.8% (Figure 1—figure supplement 1b, Supplementary File 2). Clustering based on pairwise MCS > 35% revealed only four distinct cluster heads and a limited number of outliers (Fig. 1c, Supplementary File 2, Figure 1—figure supplement 2). To quantify the potential for model performance to be driven by memorization, for each ligand in our set, we retrieved the most similar ligand that was deposited prior to the training cutoff date by both MCS and Tc (Fig. 1d). Here, the Tcs have a trimodal distribution with a small number having Tc > 0.4 (indicating relatively high similarity for this fingerprint), and two larger distributions centered around 0.35 (indicating scaffold hops) and 0.2 (indicating low similarity to any ligand). The MCS similarities spanned a broad range centered around ∼60%, consistent with the majority of the dataset representing a small number of conserved scaffolds bearing diverse peripheral substitutions. We could also confirm the chemical diversity of the Mac1 compounds relative to the broader chemical space represented by all ligands deposited in the PDB by both Tc and MCS% (Figure 1—figure supplement 1c-d).

Co-folding can accurately reproduce poses of ligands dissimilar to those trained

Next we assessed how well co-folding methods could predict the structures of the 551 complexes. Three co-folding methods (AlphaFold3, Chai-1, and Boltz-2) were given the SMILES strings of each ligand and the sequence of the enzyme as inputs. In parallel, we performed classic docking with DOCK3.7 using the SMILES of the ligand and the three dimensional structure of the enzyme bound to an early inhibitor (PDB: 5SQW, (Gahbauer et al. 2023) that is roughly contemporaneous with the ligands in this datasets. However, 68 structures had weak density for part of the molecule, and were not used as reference structures for comparing co-folded poses. To assess whether ligands were correctly predicted in the right area of the active site (Nittinger et al. 2025), we used a threshold of ligand center of mass < 2.5 Å after alignment of 489 protein structures: 449 AlphaFold3 poses (92%), 440Chai-1 poses (90%) and 444Boltz-2 poses (93%). With DOCK3.7, only 80% (392 out of 489 poses) passed this cutoff. In addition, roughly 1.5% of co-folded or docked molecules did not pass the RDkitbased sanitization pipeline necessary for RMSD comparisons. Therefore, the total number of poses compared for prospective prediction differed slightly among all co-folding and docking methods (Supplementary File 3).

Across the co-folding methods, AF3 and Chai-1 performed qualitatively similar with a roughly unimodal distribution centered around ∼1 Å (the mode) RMSD values (Fig. 2a). Using the field standard, 2 Å cutoff for correct predictions, AF3 and Chai-1 predicted roughly70% of the ligands correctly. Boltz-2 performed less well, with more structures shifting to the 2 Å distribution, and an overall predictive accuracy of 53%. This result was surprising because the later training cut-off date for Boltz-2 meant that it could potentially benefit from additional Mac1 ligand-bound structures. The distribution for docking was more bimodal centered around an RMSD of 2 Å and 5 Å, yielding a predictive accuracy of 40%. At least for this set, the co-folding models outperformed DOCK3.7 in accurately predicting the poses of bound ligands.

Co-folding outperforms DOCK3.7 in predicting the ligand pose for 489 ligands against Mac1 binding site.

a) Histogram distributions of ligand heavy-atom RMSD (L-RMSD) between predicted and crystallographic poses obtained with AF3, Chai-1, Boltz-2 and DOCK3.7, and the success rate defined as RMSD < 2 Å, b) Demonstration of alternate conformations of Mac1 binding pocket (twisted backbone and open state) and ability of AF3 to capture these conformational changes (residues in green are from PDB ID: 5SQW, and residues in pink have altered conformations), c) AF3 L-RMSD pose recovery is compared with TC ECFP4 and MCS% and accurate poses for test molecules with poor similarity to the closest molecule in the training set.

Next we investigated the potential for co-folding methods to identify alternative conformations of the protein that are stabilized by ligand binding. This application is a potentially large differentiator relative to virtual screening, which generally uses a fixed structure. Although protein alignment gave generally low global C-alpha RMSDs (range 0.1 Å – 0.4 Å; Supplementary File 3), rare local and high magnitude conformational changes observed experimentally were not captured by co-folding methods. Previously published structures indicated two characteristic states with significant conformational changes: a twisted state, characterized by a peptide backbone flip around Phe156, and an open state, characterized by a ∼8 Å loop rearrangement including Gly130 and Ala129 (Fig. 2b, Figure 2—figure supplement 1, Supplementary File 4). The open state has also been observed in the human MacroD1 enzyme (PDB 2X47, (Chen et al. 2011) and PDB 6LH4, (Yang et al. 2020)), suggesting it might be a common feature across macrodomains. We observed that only 4/19 twisted states and 0/20 open states were correctly predicted by AF3. Interestingly, AF3 predicted 8 false positive twisted states, where the ligand was experimentally observed in the normal ground state. This result suggests that there is only a limited ability to predict conformational adaptations, which admittedly may be, at least in part, stabilized by crystal contacts. We next asked how often the predictions recapitulated hydrogen bonding interactions in the Mac1 active site (Figure 2—figure supplement 1a). Among the dominant interactions, we found a low false positive rate for hydrogen bonds with D22 and I23 at the top of the active site, and mispredictions in both directions for hydrogen bonds with F156 and D157 at the oxyanion site (Figure 2—figure supplement 1b). Thus, despite failing to predict large conformational changes and aspects of the detailed hydrogen bonding interactions, AF3 still correctly placed these ligands with low RMSDs relative to the crystal structure pose (Errington et al. 2025). This result suggests that co-folding reliably recapitulates dominant ligand-binding interactions even in the absence of accurate protein conformational modeling, providing further support to the idea that they are learning specific interaction patterns rather than a deeper physics-based representation (Masters et al. 2025).Finally, we asked how much of co-folding success might be explained by memorization of Mac1 ligands deposited prior to the training date (Fig. 1d). We measured the correlation between AF3 ligand RMSD, and either the Tc (ECFP4) or %MCS of the most similar ligand within the training set, finding no meaningful signal (Fig. 2c). Examples illustrated have low similarity in both metrics, and mac-x5015, mac-x3538 are not from the commonly shared purine scaffold. Nonetheless, even for these cases, AF3 co-folded poses are recovered to within 1 Å accuracy. The weak correlations between RMSD and both similarity metrics illustrate the ability to predict binding modes for ligands dissimilar from the training set.

Model confidence and energy scores reflect pose reproduction and affinities

A central question for any co-folding or docking method is whether its internal confidence or energy scores track with experimental pose accuracy and affinity. We used the structure confidence scores reported by AF3, the per atom Predicted Local Distance Difference Test (pLDDT) scores for ligand atoms, and for Chai-1 the interface predictive template modeling (ipTM) score. While these metrics are intended to estimate the model confidence in accuracy, they are not explicitly designed to predict affinity. In contrast, the Boltz-2 affinity module predicts IC50 (pIC50) which should correlate with affinity, whilst the DOCK3.7 score, though in kcal/mol, is not expected to correlate strongly with experimental affinity (Shoichet 2004; Irwin and Shoichet 2016; Tirado-Rives and Jorgensen 2006). Boltz-2 affinity score and DOCK energies do not explicitly estimate the accuracy (RMSD) of the prediction.

We first generated Receiver Operating Characteristic (ROC) curves to measure the retrieval of True Positives (good score, RMSD < 2Å) versus False Positives (good score, RMSD > 2Å) (Fig. 3a). We observed the best performance for AF3 L-pLDDT (AUC=75.6) and Chai-1 ipTM (AUC=74.6). This indicates that the internal model confidence has predictive power and, indeed, we observed that these scores are correlated with RMSD (Figure 3—figure supplement 1). Perhaps not surprisingly, the performance of the DOCK3.7 energy (AUC=67.1) and Boltz-2 pIC50 (AUC=53.8) scores, which are designed to predict affinity, was poorer than those designed to predict accuracy. The DOCK3.7 scores correlated with Mac1 ligand RMSD with a Pearson R of 0.33(p < 0.001), which is in line with expectations from historical analyses of docking studies (Wierbowski et al. 2020). However, Boltz-2 affinity scores did not correlate with the predicted pose (Pearson r = −0.001, p > 0.5), which likely results from the fact that its internal structure prediction confidence module is separated from the affinity module (Passaro et al. 2025). Indeed, using the Boltz-2 structure prediction confidence metric (Ligand ipTM), yields an AUC of 75.2, comparable to AF3 and Chai-1. Considering its early enrichment performance, Boltz-2 Ligand ipTM was the strongest predictor of pose accuracy based on normalized logAUC (27.1% above random, Fig. 3a). In contrast, although Boltz-2 pIC50 showed poor overall discrimination, it overestimated its ability to enrich true positive poses at low false positive rates, despite having a weak early enrichment behavior. Since confidence scores correlated with RMSD, we also investigated how each method was able to predict poses and scores depending on similarity of molecules to the training set (Figure 3—figure supplement 2). As before (Fig. 2c), the performance of co-folding and docking scores did not show a strong dependence on novelty relative to the training set. This result suggests that co-folding can accurately reproduce poses of ligands dissimilar to those on which it trained, and that the model internal confidence scores reflect how well poses are reproduced.

Complementarity between co-folding confidence metrics and ligand-pose accuracy.

a) Receiver Operating Characteristic (ROC) curves for affinity/energy and confidence scores in AlphaFold3, Chai-1 and Boltz-2, with the success rate of binding pose prediction being defined as L-RMSD < 2 Åand both AUC and logAUC are shown in percentages. b) Chemical structure of a compound that was challenging to model with all docking, co-folding methods (mac-x4091), and one that was well-modelled with all methods (mac-x3927), c) Overlay of co-folded (AlphaFold3, Chai-1, Boltz-2) and docked lig- and poses on ground-truth, for hits with similar IC50, but different structures. Hydrogen bonds with co-folded, docked ligands are shown in red, and those with crystal ligands are shown in yellow.

Representative examples of correct and incorrect pose predictions across co-folding and docking methods are illustrated with mac-x3927 and mac-x4091, respectively (Fig. 3b,c). Both compounds exhibited comparable experimental IC50 values (20 µM for mac-x4091 and 42 µM for mac-x3927) and similarity to molecules in the training set (Fig. 3b). The molecule consistently predicted correctly (mac-x3927) was a small, rigid fragment comprising a purine scaffold rigidly linked to a fluorobenzene moiety, favoring a well-defined binding mode. In contrast, the compound mispredicted by all methods (mac-x4091) was larger and more conformationally flexible, with an alkyne linker prone to flipping, potentially altering the orientation of the fluorobenzyl ether substituent. Although this group packs against a crystal contact, the high predictive confidence/affinity scores across co-folding methods (AF3 L-pLDDT = 86.4, Chai-1 ipTM = 0.91) are likely explained by the adenosine mimic purine group where predictions recapitulate key interactions observed in the experimental structure.

To understand if prediction errors between methods are correlated, we directly compared the agreement between AF3 to Chai-1, Boltz-2 or DOCK3.7 predictions with the agreement between AF3 and experimental ground truth (Figure 3—figure supplement 3). We observed that while there are a small number of structures where AF3 predicts much better than other methods (high AF3-other prediction RMSD, low AF3-experiment RMSD), most large errors are either missed in a similar way (low AF3-other prediction RMSD, but high AF3-experiment RMSD) or are both extreme mispredictions (high AF3-other prediction RMSD and AF3-experiment RMSD). Consistent with this result, in comparing RMSD relative to experimental structures for all methods (Figure 3—figure supplement 4), we observed that Chai-1 and AF3 exhibited the strongest agreement (r = 0.72), whereas poses from DOCK and Boltz-2 diverged more (r = 0.45 for AF3 vs. DOCK, and r = 0.52 for AF3 vs. Boltz-2). Collectively, these results suggest that the co-folding methods (AF3, Chai-1) may be complementary to more physics-based approaches like DOCK3.7 as the errors are less correlated than observed for the different cofolding methods.

Second, we asked how well these scores correlate with experimentally measured potency (Fig. 4a). The correlation was best for the Boltz-2 pIC50 (r = 0.6), significantly outperforming DOCK when tested with Friedman and Conover tests with Bonferroni corrections (p = 0.0057). Prior to any linear correction, the mean absolute error (MAE) in affinity prediction was ∼1.6 kcals (1.19 pIC units), with most affinities predicted as too potent. Empirically correcting with a linear fit reduced this error to ∼0.7 kcals (0.54 pIC units). To put these results in context, simply predicting the mean pIC50 from the dataset for every compound would yield a baseline MAE of 0.93 kcal because of the relatively limited dynamic range of the dataset. The DOCK3.7 energy correlated more weakly with potency (r = −0.225), consistent with previous studies that have observed only weak correlations with affinity from virtual screening energies (Kinnings et al. 2011; Jones et al. 2021). The AF3 co-folding confidence metrics correlated slightly better than docking with potency (r = 0.314, Fig. 4a), but the post-hoc Conover tests with Bonferroni corrections showed the difference was not statistically meaningful. Despite this, a correlation between pose accuracy and potency was more relevant with AF3 (r = −0.233, Fig. 4b), as more accurate poses tend to be more confidently predicted (Figure 3—figure supplement 1). In contrast, there was no correlation between docking accuracy and potency (Fig. 4b). Boltz pIC50 had the highest correlation between pose accuracy and affinity (r = −0.279). This result is interesting because in Boltz-2 affinity prediction and confidence are decoupled architecturally and have little correlation with each other across the Mac1 dataset (Figure 3—figure supplement 1).

Co-folding scores and pose accuracy correlate with Mac1 binding affinity for 202 newly synthesized Mac1 ligands.

a) AF3 ligand pLDDT (L-pLDDT) and Boltz-2 predicted pIC50 correlate with experimental pIC50 more strongly than DOCK energy scores, b) Lower co-folded ligand RMSDs associate with higher affinity, while docked pose RMSDs do not. Pearson’s Correlation Coefficients and Mean Absolute Errors (MAE) are shown. The baseline MAE was 0.68 pIC (or 0.93 kcals), measured against a fixed average pIC50 (pIC50 = 4.95). Friedman tests with Conover post-hoc comparisons and Bonferroni corrections are used to compare MAE between methods.

Collectively, these results suggest that the categorization abilities of the co-folding confidence metrics, and their correlation with affinity, could mark an important advance. Still, the Mac1 set, as large as it is, represents only a single system with ligands that share a limited set of pharmacophores. We wanted to broaden testing to more targets and more diverse ligands, such as those encountered in virtual screening hit lists with their focus on new chemotype discovery.

Co-folding struggles to reliably re-rank Docking Hit-lists

Encouraged by the ability of co-folding to reproduce ligand structures and affinities, we were interested to see if the methods could also improve hit identification from large library docking screens. In these screens, hundreds of millions to billions of molecules are docked against a target and scored for complementarity to a binding site. High ranking molecules are then synthesized and tested for binding. While large library docking has successfully discovered potent novel chemotypes in recent studies (Lyu et al. 2019; Gorgulla et al. 2020; Stein et al. 2020; Sadybekov et al. 2022; Fink et al. 2022; Singh et al. 2023; Liu et al. 2025), the method is plagued by false positives (molecules that rank well but are inactive). We reasoned that if co-folding can predict the structures and activities of true ligands confidently, perhaps it could distinguish docking true binders from false positives.

To test this idea, we analyzed experimentally tested compounds from three large and heavily tested docking campaigns (Supplementary File 5): 506 experimentally-tested compounds from a 490-million molecule screen against the σ2 receptor (Alon et al. 2021), 1,293 tested compounds from a 1.7-billion molecule screen against AmpC β-lactamase (Liu et al. 2025), and 541 tested compounds from a 138-million molecule screen against dopamine D4 receptors (Lyu et al. 2019) with AF3 (the best performing model from the Mac1 set), DOCK scoring function, and the Boltz-2 pIC50 score. These test sets differ from the Mac1 benchmark in three ways. First, they are a mix of molecules that would turn out to be true binders (247 in AmpC, 201 in σ2 and 205 in Dopamine D4) and high-ranking docking false positives as judged by measured potencies (worse than 400 μM in AmpC, worse than 2.0 μM against the σ2 receptor, and worse than 7.8 μM against the dopamine D4 receptor). For these molecules, the challenge is not only predicting the bound structures of true ligands, but distinguishing true ligands from decoy molecules that by many biophysical criteria fit the binding site well. Second, the docking molecules are more diverse than the Mac1 set (Figure 1—figure supplement 2), which largely falls into four substructure clusters representing conserved pharmacophores. The docking sets were selected, in the original studies, for their topological diversity, and their dissimilarity to previously known ligands. For each target, the ligands group into more than 10-times the number of clusters than the Mac1 dataset, consistent with a greater diversity. Indeed, the average pairwise ECFP4 Tc values within the three sets vary from 0.12 to 0.15, all well-below scaffold-hop cut-offs for this fingerprint (Figure 1—figure supplement 2, Supplementary File 2). Finally, the sets were all selected for novelty and there was little similarity between the ligands in the hit-lists and known ligands, which ranged between 0.11 and 0.19 by Tc, and between 33 to 42 by MCS%. These ranges were lower than our prior Mac1 dataset (where Tc is distributed around 0.36, and MCS% is distributed around 60.1%), and the lack of similarity with the training set for target systems with hit-lists could be explained by having relatively fewer ligand structures to compare to in the PDB (Figure 4—figure supplement 1, Supplementary File 6). These conditions, while representing new and difficult tests for co-folding, are the common case for structure-based ligand discovery campaigns. For validation, there are more measured IC50 values, but they are admittedly not as deeply supported by structural information as the Mac1 set.

We observed the strongest discrimination between active and inactive molecules in the σ2 dataset, which contains 201 confirmed actives (apparent Kivalues between 2.5 nM and 1.5 µM) and 305 inactive false positives (Fig. 5a). DOCK3.7 scores well-separated binders from non-binders (median –55.0 vs –42.5 kcal mol-1; AUC = 78.8), though the correlation between docking score and pKi for true ligands was relatively weak (r = –0.211). AF3 L-pLDDT showed a poorer ability to discriminate between true ligands and false positives (median 75.9 vs 74.1, AUC = 56.1). Intriguingly, despite the worse discriminatory ability, the AF3 correlation with binders pKi was not much worse than docking’s (r = 0.183). Against the σ2 set, Boltz-2 pIC50 performed better still, with high discriminatory power (median 6.70 vs 5.38, AUC = 83.8) and the strongest correlation of affinity predictions against true positives (r = 0.486). Among active molecules, Boltz-2 pIC50 had a MAE of only 0.7 kcal and a linear calibration reduced that error to 0.54 kcal (Figure 5—figure supplement 1). This absolute estimation of IC50 was superior to its performance by ∼0.5 kcal on the Mac1 dataset; however, it is restricted only to true positives and the performance could be much worse if the false positives were accounted for in the comparison. Since inactive predictions were in the same pIC50 range as active molecules, as an optimistic limit of the performance that includes inactive molecules, we assigned an IC50 of 4 µM, which is 2X our threshold for declaring a non–binder, to all false positives and calculated the MAE. With this assumption, the MAE is 0.92 kcal (0.56 kcal after linear calibration). A less generous assumption would be to randomly assign an IC50 between the pKi threshold (2 µM) and pKi = 1 (100 mM). Under this assumption, the MAE is 1.66 kcal (1.55 kcal after linear calibration) (Figure 5—figure supplement 2), which is in line with previously published estimates of the accuracy of this method (Passaro et al. 2025) and only a slight improvement over the MAE (1.84 kcal) produced by predicting the dataset average affinity for each datapoint.

Classification of known active hits from docked false positives from an initially-screened docked molecules.

The leftmost panel shows affinity scores from Boltz-2 (Boltz-2 pIC50), the middle panel for AF3 L-pLDDT and the right panel for absolute value of DOCK energies (kcal/mol) from DOCK3.7. Docked hit lists from a) σ2, b) Dopamine D4, c) AmpC are grouped into known hits and non-hits, and score distributions are shown in violin plots. Komolgorov-Smirnov test and its KS-statistic (KS stat) is used to indicate meaningful separation of two distribution curves. Asterisks indicate the p-value of KS statistic < 0.001. ROC curves show an ability to distinguish known actives from the docked list with different scoring methods.

For AmpC (Fig. 5b), a total of 1,293 high-ranking molecules were experimentally tested for inhibition (Liu et al. 2025). Among these were 247 molecules that turned out to be true inhibitors. We observed that docking (AUC = 76.5, median −85.6 vs −82.8 kcal/mol) was much stronger than Boltz-2 (AUC = 68.1, median 5.16 vs 4.87) or AF3 (AUC = 60.5, median 76.6 vs 72.0) in discriminating actives from inactives. The correlation of metrics to pKiamong actives was relatively weak for DOCK3.7, AF3 and Boltz-2 in this case. The Boltz-2 pIC50 values for AmpC were particularly high in absolute error 1.81 kcal; however, linear correction reduced this error to only 0.42 kcal. Including inactive molecules degraded performance to 2.33 kcal (0.32 kcal with linear correction) under the optimistic and 3.29 kcal (1.04 kcal with linear correction) under the random IC50 assumptions (Figure 5—figure supplement 2). Similarly, this represents a small reduction relative to the MAE (1.07 kcal) produced by predicting the dataset average affinity for each datapoint

The dopamine receptor also proved to be more challenging than σ2, particularly for AF3. With D4 with 205 confirmed actives (apparent Ki values between 3.1 nM and 7.8 µM) and 336 inactive false positives with, we observed that actives were distinguished from docked false positives by docking (−63.9 vs −52.0 kcal/mol) and Boltz-2 (6.15 vs 5.72) scores, both with AUC ∼ 71. In contrast, AF3 L-pLDDT values showed little separation (65.4 vs 66.2) and an AUC = 46.4, which is slightly worse than random (Fig. 5c). There was also no clear evidence of early ligand enrichment with co-folding methods over docking methods from semilogarithmic ROC plots across the three virtual screen targets (Figure 5—figure supplement 3). Among active molecules, the docking correlation was slightly stronger for D4 than it was for σ2 (r = −0.247), but performance degraded for AF3 (r = −0.109, notably a positive correlation is expected) and Boltz-2 pIC50 (r = 0.284). Despite this reduced linear correlation for the Boltz-2 pIC50, the MAE was 0.89 kcal (0.55 kcal after correction). The performance degraded to 1.15 kcal (0.57 kcal after linear correction) under the optimistic assumption of 2X IC50 for all false positives and to 1.97 kcal (1.4 kcal after linear correction) under the randomly assigned assumption (Figure 5—figure supplement 2). This result is also a small improvement over the MAE (1.46 kcal) produced by predicting the dataset average affinity for each datapoint

An alternative way of looking at how well the scores can identify active ligands is to perform a hit enrichment analysis using normalized ranking scores (pProp, the negative logarithm of the percentage rank (Liu et al. 2025): a molecule that ranks among the top 0.1% of the docked library would have a pProp of 3, one that ranked among the top 0.01% of the docked library would have a pProp of 4, and so forth). pProp assigned here is based on the ranking within the docked list, for direct comparison between scoring methods and largely confirmed the trends above (Figure 5—figure supplement 4). Both DOCK3.7 and Boltz-2 clearly enriched actives across most targets, with hit rates plateauing at 40 - 50% for high-ranking compounds. AF3 L-pLDDT showed meaningful and comparable enrichment only for AmpC, while providing no enrichment signal for σ2 and D4. Overall, the success of the co-folding methods at predicting structures and correlating with affinities among true Mac1 inhibitors was far less apparent in their ability to distinguish true-from false-positives among docking hit-lists, and to correlate with affinities among these highly diverse molecules. Indeed, the methods showed little advantage over docking, a method that is about four orders of magnitude faster. An important caveat is that the hit-lists were composed of molecules prioritized by docking in the first place, giving it an advantage on these particular sets.

Discussion

A striking observation from this study was the ability of the co-folding methods to predict ligand structures across 551 previously unpublished Mac1 complexes, and for the co-folding confidence metrics to correlate with ligand affinities. By both criteria all co-folding methods were meaningfully better than docking, at least as represented by DOCK3.7. Affinity prediction by Boltz-2 also performed well with mean absolute error that was ∼0.2kcal better than a baseline expectation of simply using the mean IC50 value. This small improvement was a trend that also held up in the large docking sets as well. Conversely, when confronted with high-anking molecules from large-library docking, a diverse mix of true- and false-positives with thousands of novel molecules prospectively tested against three unrelated targets, co-folding was not meaningfully better than docking at ligand categorization or at affinity prediction, indeed often worse. This despite the four log-orders greater speed of the docking and its well-known approximations and errors, long a topic of regret in the community (Jorgensen and Tirado-Rives 2005; Irwin and Shoichet 2016). These apparently contradictory results may be reconciled by the different tasks of predicting geometries for molecules that were all true ligands versus that of categorizing binders and non-binders within a highly diverse docking hit-list intentionally dissimilar to known ligands.

It may be that success and relative failure in these different tasks points also to a complementarity between physics-based docking and co-folding in the related, but different tasks of ligand discovery and optimization. The weak correlation between the methods (Figure 5—figure supplement 5) further suggests complementarity between deep learning and physics based approaches. Docking and related methods are designed for categorization among diverse, novel sets of candidate molecules, drawn from the vast new make-on-demand space (Gentile et al. 2020; Radaeva et al. 2023; Sadybekov et al. 2022; Sadybekov and Katritch 2023; Gorgulla et al. 2021; Lyu et al. 2019, 2023; Liu et al. 2024, 2025), whereas the co-folding methods, relying on learned patterns, may be relatively less effective. Where they have complementary strengths is in predicting the accurate geometries of ligands, once known, and their ranking within a series, both of which are crucial for affinity maturation once a series is discovered and with both of which docking has historically struggled. It is in these two linked but different aspects of ligand discovery and ligand optimization where the true complementarity between the methods may lie, at least for now.

Several caveats merit airing. The Mac1 benchmark, while diverse by ECPP4 Tc, is nevertheless dominated by four groups when clustered by MCS and benefits from a large training set of fragment molecules bound to the protein in the training set (Schuller et al. 2021). This may inflate apparent pose recovery rates, although Boltz-2 predicted poses were poorly recapitulated even with its much later training cutoff dates. Unexpectedly, even when the co-folding methods got important conformations of the enzyme binding site wrong, the ligands were still often posed correctly, speaking to a non-physical aspect of the ML methods that others have also described (Masters et al. 2025; Škrinjar et al. 2025). Even though the prospective docking hit lists included thousands of diverse molecules, they still represent only three targets, and so offer a limited view of molecular recognition. Finally, comparing co-folding to docking based on hit-lists themselves selected by docking is arguably unfair to co-folding. Counterbalancing this is the inclusion, in each of the three hit lists, of molecules that had mediocre and poor docking scores intentionally selected to test the correlation between docking score and hit-rate. Here too, the correlation between co-folding score and likelihood to bind, what we sometimes call a “dock-response-curve” was no better than docking’s, often worse (Figure 5—figure supplement 4).

These caveats should not obscure the central lessons from this study. Against a benchmark of 551 topologically diverse Mac1 ligand-bound structures, determined to true atomic resolution and largely dissimilar to structures already in the PDB, each of the co-folding methods predicted ligand poses and interactions with high-fidelity to the experimental structure. For Boltz-2, there was a strong and highly significant correlation between predicted affinity and experimental IC50 (compared to a modest correlation for AF3 L-pLDDT and experimental IC50). In both cases, co-folding performance was much better than that of DOCK3.7, although the correlation was only significantly different for Boltz-2 pIC50 versus experimental pIC50. Conversely, when the more homogeneous benchmark of true ligands was replaced by a diverse set of docked molecules, composed of what would turn out to be hundreds of true ligands and thousands of false positives, the co-folding methods did no better than docking and often did worse. These observations are supported by a fascinating study on some of the same ligand sets as investigated here, using AlphaFold3, reaching similar conclusions (Menon et al. 2025). The relatively strong performance by docking is surprising given its well-known weaknesses and its much reduced computational cost (and therefore greater speed). What emerges are approaches that may be complementary in their roles in ligand discovery (docking) and structure-based ligand optimization (co-folding), areas that will merit future research and application.

Methods

Mac1 X-ray crystallography

Ligands were soaked into SARS-CoV-2 Mac1 crystals (P43 crystal form (Schuller et al. 2021)), and data collected and reduced, as described previously (Gahbauer et al. 2023; Correy et al. 2025). We modeled ligands into PanDDA event maps (Pearce et al. 2017) using ligand coordinates and restraints generated with phenix.elbow (Moriarty et al. 2009).

Ligand Similarity, Chemical Diversity and Memorization Analyses

To test whether pose-prediction success reflected latent memorization, we quantified chemical similarity between test and training ligands using two complementary metrics. First, extendedconnectivity fingerprints (ECFP4) were computed for all molecules, and Tanimoto coefficients (TC) were used to identify, for each test ligand, the most similar compound present in the training corpus (Chai-1 entries deposited before 1 December 2021 and AlphaFold3 entries before 30 September 2021). The training date cutoff for Boltz-2 is 1 June 2023 (Supplementary File 6).

As another measure of structural similarity, the maximum common substructure percentage (MCS%) was calculated using the rdFMCS algorithm in RDKit, reporting the proportion of heavy atoms shared between two ligands. To evaluate chemical diversity within benchmark datasets, pairwise TC or MCS% values were computed across all ligands. Best-First Clustering (BFC) based on these similarity metrics was used to estimate the number of unique chemotypes represented in each dataset. Principal component analysis (PCA) was used to visualize the resulting chemical space, and to identify an appropriate threshold for MCS-based scaffold grouping. Within the Mac1 benchmark set, applying an MCS cutoff of 35% yielded four distinct scaffolds, in agreement with PCA-derived cluster separation. The same clustering thresh-old was subsequently applied to define scaffold families for the AmpC, σ2 and D4 benchmark datasets.

Molecular Docking of Mac1 compounds

The modelled orthosteric site of the Mac1 model was available from the PanDDA analysis group deposition in the Protein Data Bank (PDB), with PDB ID: 5SQW (Gahbauer et al. 2023). We used this structure because the inhibitor (Z5014193706) was the most potent molecule with a structure determined around the same time as the ligands in this dataset were tested. This structure was used to generate matching spheres, which are later used by DOCK3.7 to pre-sample ligand conformations into binding sites by superimposing ligand atoms on spheres in the site. 45 matching spheres were used, and the docked ligands were scored by summing receptor-ligand electrostatic and van der Waals energies and corrected for context-dependent ligand desolvation. Receptor structures were protonated using Reduce. Partial charges from the united-atom AMBER force field were used for all receptor atoms. Potential energy grids for the different energy terms of the scoring function were precalculated using AMBER for the van der Waals term and the Poisson-Boltzmann method QNIFFT for electrostatics. Context-dependent ligand desolvation was calculated from the Generalized Born method. Ligands were protonated at pH7.4 (ChemAxon), and each protomer was rendered into 3D using Corina, followed by conformational sampling using Omega. For the docking campaign of 551 novel compounds, each molecule was sampled in about 2,719 orientations and, on average, 105 conformations by. Overall, over 290 million complexes were sampled and scored.

Co-folding

All three co-folding methods (AlphaFold3, Chai-1 and Boltz-2) were available from following sources:

- AlphaFold3: https://github.com/google-deepmind/alphafold3

- Chai-1: https://github.com/chaidiscovery/chai-lab.git

- Boltz-2: https://github.com/jwohlwend/boltz

Ligand SMILES and amino acid sequences for the target protein were used to perform co-folding in an array of NVIDIA A40 GPUs. Multiple Sequence Alignment (MSA) was performed with the Jackhammer module. Co-folding tools will generate predicted protein-ligand complex poses (mmCIF files), and different confidence metrics for these predictions. Chai-1 and AlphaFold3 generated global confidence scores (pTM, ipTM), whilst AlphaFold3 had local atom-specific metrics, such as, pLDDT, PAE and contact probabilities in large vectors. To put an emphasis on prediction of ligands, we computed a ligand-centric pLDDT (L-pLDDT) by taking an arithmetic mean of pLDDT on ligand atoms, and a ligand-centric PAE (L-PAE) by taking an average of all PAE elements that contain ligand atoms. Minimum PAE (mPAE) was obtained by scanning all PAE elements that involve ligand atoms, and finding the smallest error. Boltz-2 generated all the confidence metrics in Chai-1 and AlphaFold3, but its affinity module generated an output of the binding likelihood (probability metric) and the predicted binding affinity (Boltz-2 pIC50) of compounds. Scripts to prepare input .json, .fasta files, postprocess output files, are available in https://github.com/jongbin99/Cofolding.

Benchmarking Co-folding on Experimentally Validated Compounds

With novel compounds discovered by fragment-based screening, 551 Mac1 complexes were resolved by the crystallography work. The same sequences were used to perform co-folding with AlphaFold3, Chai-1 and Boltz-2. For further sets of analyses with docked hit lists on targets from the Large-Scale Docking (source: lsd.docking.org)(Hall et al. 2025), only AlphaFold3 and Boltz-2 were used. For each system, we redefined the threshold for true hits and non-binders as described herein:

AmpC β-lactamase: Among 1.7 billion docked molecules, the top scoring 1% were selected (n = 1,293) (Liu et al. 2025). We selected ones with apparent Ki < 400 μM (n = 247) as true hits, and the rest were regarded as non-binders.

σ2: True hits among these docked molecules were previously defined as displacing greater than 50% of 3H ditolylguanidine ([3H]DTG) to σ2 (Alon et al. 2021), but in our work, the cutoff is adjusted to 25% displacement, giving a less stringent threshold to include more molecules into the space of true hits. Single concentration point measurement (at 1 μM) is converted to an apparent Ki value using Cheng-Prusoff equation (Ki=IC501+[S]Km), with pre-calculated values of [S]Kmbased on radioligand displacement (Alon et al. 2021). This yields a non-binding cutoff of an apparent Ki < 2.0 μM.

Dopamine D4: For the dopamine receptor D4, 138 million molecules were docked and then prioritized synthesizable high-ranking docked molecules (n = 549). True hits from these molecules were defined as displacing more than 50% of 3H-N-methylspiperone (Lyu et al. 2019), but similarly to σ2, we adjusted the cutoff to 25% displacement. Apparent Ki values are calculated similarly to σ2 case, but at 10 μM. This yields a non-binding cutoff of an apparent Ki < 7.8 μM.

Ligand Pose Evaluation via RMSD against crystal structures

Predicted ligand binding modes were benchmarked against experimentally determined reference poses from ultra-high-resolution (∼1.0 Å) crystal structures. The Mac1 protein receptor was first superimposed by least-squares fitting of all Cα atoms using PyMol. The heavy-atom root-mean-square deviation (RMSD) between the aligned co-folded ligand and its crystallo-graphic counterpart was then calculated. Before RMSD evaluation, both predicted and reference ligands were confirmed using a uniform RDKit-based sanitization pipeline (Buttenschoen et al. 2024). SMILES strings were converted to RDKit Mol objects, chemically validated (valence, aromaticity), and assigned bond orders from a canonical SMILES-derived template. We then check that ground-truth (experimental) and co-folded ligand coordinates for all molecules to have valid chemistry as described with the SMILES-derived template. A substructure match then enabled a one-to-one correspondence of heavy atoms between the ground-truth and co-folded poses for comparison. The crystal ligand and co-folded ligand had the center-of-mass (COM) calculated separately, and measured the distance between each COM. If the distance exceeds 2.5 Å, we would expect the RMSD to be large, indicating that lig- and poses are not modelled accurately in an orthosteric pocket (Nittinger et al. 2025). For all the predicted ligand poses, rdMolAlign.CalcRMS from Rdkit.Chem is used for calculating the ligand RMSD between co-folding and the ground-truth crystal pose.

Hit-rate curves for different scoring methods

To get hit-rate curves for co-folding and docking scores across three targets of docked hitlists (AmpC, σ2, D4), molecules in the list are ordered by increasing score (for DOCK) and decreasing score (for L-pLDDT and Boltz-2 pIC50). A rolling window of 100 was used for AmpC and σ2, and 50 for D4 (Liu et al. 2025), calculating the hit rate as a fraction of molecules with experimental binding affinity that match our definition for true hits described above (refer to Benchmarking Co-Folding on Experimentally Validated Compounds in Method section). The docked hit-list is a mixture of known actives and non-binders that include false positives based on docking scores. For direct comparison of hit classification performance between DOCK energies, AF3 L-pLDDT and Boltz-2 pIC50, the pProp (the negative base 10 logarithm of the fractional rank) was assigned within the list, instead of the billion-scale full docking list. For each window, we used a 95% Wilson confidence interval with normal approximation for the hit rate.

Statistical Analysis

Pearson’s correlation coefficients were used throughout the paper to quantify linear associations between two variables in the study. We have also reported a mean absolute error of data points with respect to the regression line for Boltz-2 affinity pIC50 against the measured pIC50. The mean absolute errors are also calculated for the perfect fit (Pearson r = 1) and for the fixed measured pIC50 (Pearson r = 0), to give an indication of how accurate the Boltz-2 scores are with respect to actual binding energies (in kcal), in terms of correlations and absolute errors. Friedman and post-hoc Conover tests were used to compare how each of AF3, Boltz-2 and DOCK3.7 score predicts for actual binding affinity, and Holm-Bonferroni correction for pairwise comparison p-values (Ash et al. 2025). The distribution of scores for known actives and docked false positives were compared using the Kolmogorov-Smirnov (KS) non-parametric statistical test, as shown in violin plots, and the KS statistic value indicates how distant the two distributions are, and p-values show the extent of statistically detectable differences. All analyses were performed using Python packages.